### 2016-05-13 paper.tex
initial commit

\documentclass[a4paper, 11pt, titlepage]{report}

\usepackage[margin=2.0cm]{geometry}

\usepackage{xspace}
\def \iZc {I\textsuperscript{2}C\xspace}
\def \vdd {V\textsubscript{DD}}

% Indent the first line of the first paragraph like the first lines of subsequent paragraphs
\usepackage{indentfirst}

% Don't indent any first lines
\setlength{\parindent}{0mm}

% A little space between paragraphs
\setlength{\parskip}{1.5mm}

% Display
% 1. <chapter title>
% instead of
% Chapter 1
% <chapter title>
\makeatletter
\renewcommand{\@makechapterhead}[1]{%
\vspace*{50 pt}%
{\setlength{\parindent}{0pt} \raggedright \normalfont
\bfseries\Huge\thechapter.\ #1
\par\nobreak\vspace{40 pt}}}
\makeatother

% In order to include images
\usepackage{graphicx}

% Label figures 1, 2, 3, 4... Instead of 1.1, 1.2, 2.1, 3.1 etc.
\usepackage{chngcntr}
\counterwithout{figure}{chapter}

\usepackage{siunitx}

% To have text wrapped around images 
\usepackage{wrapfig}

% Some parameters for wrapfig
\setlength{\intextsep}{0pt}%
\setlength{\columnsep}{15pt}%

% Same as regular math but with a bit of space before and after
\newenvironment{ownmath}
{\vspace{2mm}\hspace{15pt}\begin{math}}
{\end{math}\vspace{2mm}}

\begin{document}

\title{Development of a 3D display}
\author{Balduin Dettling}

% Generate title
\maketitle

% Introduction
% * after chapters to not number them
% TODO: don't show page numbers
\chapter*{Preface}

\pagenumbering{Roman}

\section*{Motivation}

For as long as I can remember, I've been fascinated by all kinds of technology. 
It was clear to me that my final thesis would be a practical one in the field of engineering.
However, I didn't have anything specific in mind for quite some time.

One day, I stumbled upon a project report on the internet about a so-called persistence of 
vision display. It was a simple row of LEDs, driven by a microcontroller, and mounted on an 
electric motor. The contraption was (and presumably still is) able to draw images into the 
air by precisely controlling the row of LEDs. Fascinated by this, I did some research and 
found that a plethora of similar projects exists, some even with detailed instructions on how 
to build the device in question.

I had the idea of building several rows of LEDs and arranging them on top of each other. 
That way, I could display several two-dimensional bitmap images above another, resulting in a
three-dimensional image.

Having had no previsous experience in the field of electronics, it seemed a bit daunting to 
build a 3D display. I could, however, motivate myself to attempt it anyway by telling myself
that the project wouldn't be much more complex than all of the 2D versions out there (which 
turned out to be not very far from the truth).

\section*{Credits}

First and foremost, I would like to express my gratitude to Mr. Patrick Spengler, my physics
teacher and the mentor of this thesis. Not only did he solve numerous malfunctions, but he also
is partly responsible for my interest in electronics and engineering.

I am thankful to Ms. Ursula Schamberger, teacher of the woodworking class, where I built the
wooden frame. Likewise, I would like to thank Mr. Hanspeter Rieder, who kindly granted me
access to the school's electronics workshop.

Last, but most certainly not least, I am indebted to Prof. Heinz Domeisen. He mentored the
advancement and refinement of the thesis in preparation for the national competition organised
by "Swiss Youth in Science".

\section*{Declaration of authorship}

I hereby declare that the following thesis has been authored entirely by myself, Balduin
Dettling, and confirm that all external sources of information are cited correctly.

\vspace{10mm}
\noindent Location and Date: \hspace{0.35\textwidth} Signed:\\
\rule[3mm]{\textwidth}{0.5pt}

% Generate table of contents
\tableofcontents

\chapter{Introduction}

\pagenumbering{arabic}

\section{Current State of The Art}

% todo: references
As discussed before, there is no shortage of projects that are similar to mine, but only
display 2D images. The name "POV display", short for persistence of vision display, has become
commonplace. Another popular name is "Propeller Clock" --- Propeller because it rotates like one,
and clock because the round shape is ideal for displaying an analog clock.

As far as I know, however, only two people have built a working version of a three-dimensional
POV display so far (and have published it on the internet).\footnote{It turned out that there was
a third one: When I posted a picture of my project on the internet, a reddit user told me about
the display he built years before.}
Neither of these is being produced on a grand scale, instead they are --- like most 2D versions
--- side projects of makers or engineers.

\section{Goals}

The objective of this project is to build a working prototype of a 3D persistence of vision
display and to document the process of doing so. The necessary steps are (roughly) as follows:

\begin{itemize}
	\item Understand and explain how and why such a display can work in principle
	\item Come up with realistic technical specifications, founded on the previously outlined 
	theory  
	\item Select parts and build the device according to the specifications
	\item Write a program that makes the device display a 3D image
	\item Draw conclusions: Did I achieve my goals? What could have been done better? 
	How can the project be further improved?
\end{itemize}

\chapter{Theory and Planning}

\section{Principle of Operation}

\subsection{Persistence of Vision}

Persistence of Vision refers to the optical illusion that results whenever the image we look
at quickly changes. In such situations, we don't immediately see the new image, instead we
continue seeing the old image for a short timespan after it has vanished. If wikipedia is to 
be believed, this timespan is about $\frac{1}{25}$ of a second, but of course it is going to 
be different for each pair of eyes.

Although we don't normally see this effect in action, there are situations where it becomes
apparent. For example, if a car drives by at nighttime, we see traces of light behind the
actual car lights for a short moment (Assuming our eyes don't follow the car and stay still
relative to the environment). In general, bright flashes of light followed by (relative)
darkness tend to be the most pronounced manifestation of persistence of vision. It is exactly 
this situation that can be recreated and exploited for the successful operation of a persistence
of vision display.

\subsection{Exploiting Persistence of Vision}

Normally, Persistence of Vision is nothing more than an interesting effect. However, we can 
use this weakness of the human eye to our benefit.

If a light source is spun quickly enough in a circle, we don't see a spinning
point anymore, but instead we see one solid circle. This is because the rotational period is
shorter than the time it takes for the light to fade away on our retinas. The rotational period
at which we stop seeing a flickering circle and start seeing a solid, still circle is called
the flicker fusion threshold.

By controlling this light source quickly and accurately, we can make it so it is always turned 
on at certain points in the circle, and always turned off at others. By doing so, we are
effectively multiplexing a zero-dimensional display (a point-like light source, e.g. an LED)
so that it displays a one-dimensional image. By rotating a whole row of LEDs (a one-dimensional
display), we can display two-dimensional images. Finally, if there are multiple rows of LEDs
that together form a two-dimensional display, it is possible to display a three-dimensional
image.

\section{Specifications}

\begin{itemize}

\item Resolution of 16 * 100 * 10 pixels (radius * circumference * height)
\item Each pixel consists of an RGB LED
\item Colour depth of 3 bit per pixel, or one bit per LED
\item Rotational frequency of at least 30 Hz, or as fast as it needs to be in order to not flicker

\end{itemize}

\section{Components}

\subsection{Control}

For controlling the LEDs, I have bought a Teensy 3.1 microcontroller development board. It has a
small footprint --- ideal for the fast rotation --- and still packs quite some power: The processor
runs at 96 MHz at a register width of 32 bit. There are 34 configurable I/O pins.

The controller has to have some kind of reference point, so that it can calculate the current speed
and angular position of the device. For this, I used a simple hall sensor on the rotating side, 
which passes by a stationary magnet after each rotation.

\subsection{Data Transmission}

34 I/O pins are quite a lot, but not quite enough to directly control 480 LEDs. In order to still
access all those LEDs individually, I had to use LED drivers. These come in the form of integrated
circuits and receive some amount of data that they use to control a corresponding amount of LEDs.

Initially, I had planned to implement a colour depth of 8 bits, allowing for 
$2^{24} = 16777216$ different colours. However, there were several difficulties that I would have
had to deal with. First of all, all of the available LED drivers had one or both of the following
shortcomings:

\begin{itemize}

\item Controlled using \iZc  instead of SPI

Because \iZc uses pullup resistors instead of a push-pull drive, it takes a little moment for the
logic level to return to its normal state. In order to ensure correct data transmission, it is 
necessary to set a relatively slow clock speed. The two \iZc interfaces of the Teensy only support
clock speeds of 2.4 MHz --- and because \iZc requires some overhead data to send out the slave's
addresses, the actual data rate will be even lower.

However, the minimum data rate required to control the LEDs at an 8-bit colour depth is as follows:

\begin{ownmath}
% Yes it's a mess but \si makes it a unit, so it is not in cursive while still being small enough
24\,\frac{\si{bit}}{\si{pixel}} * 16000\,\frac{\si{pixels}}{\si{rotation}} * 30\,\frac{\si{rotations}}{\si{second}} = \SI{11.52}{Mbps} 
\end{ownmath}

This means that I'd need at least 5 parallel \iZc lines to control all the LEDs with 8 bit colour
depth. Because each Teensy only has two of them, I'd need three Teensy boards, which would be
possible but very cumbersome.

\item Insufficient PWM frequency

If the colour depth is more than one bit, the LEDs will have to be dimmed using PWM. Normally, a
rather low frequency of a few hundred to a few thousand Hz is sufficient as a PWM frequency, and
most LED drivers modulate at frequencies within that range. In my case, however, the LEDs are 
moving into a new pixel 3000 times a second (assuming rotationat 30 Hz). This means that a PWM
frequency under 3000 Hz will involve losing information, and one slightly above 3000 Hz still
won't look good. There would have to be a significant number of PWM cycles within each pixel,
for example 10, which would require a PWM frequency of 30 KHz.

\end{itemize}

Because of that, I ditched the idea of 8-bit colour depth and instead went for the simpler approach
of one-bit colour depth. Both problems are solved by switching to 1 bit: The data rate doesn't have
to be as high (and there are lots of chips that are controlled by SPI), and PWM is not necessary
anymore.

% todo: Find out if this is necessary in the end
\newpage
\subsection{LED drivers}

\begin{wrapfigure}{r}{0cm}

\includegraphics[scale=0.35]{./images/tlc-pins.png}
\caption{Pin layout of the TLC5927 LED driver}

\end{wrapfigure}

After quite some searching around, I decided to use the TLC5927 by Texas Instruments. This is
essentially a shift register with some extra features. Like every shift register, it has a serial
data input (SDI), a serial data output (SDO), and a clock pin (CLK). Additionally, there is a latch
pin (LE), which serves the purpose of only refreshing the state of the LEDs once all data has been
sent.

Furthermore, there is a pin called "output enable" (OE). If it is at GND, all LEDs are turned off,
and it's at \vdd, they are controlled by the data that the driver chip has received.

The most important feature, however, is the fact that each output is controlled by a constant
current driver. Without these, it would've been necessary to solder a resistor in series
with each individual LED, for a total of 480 resistors, 960 additional solder joints and a whole
lot of wasted board space. With the constant current drivers, however, I could use a single
resistor to set the current for all 16 LEDs attached to one chip.

Unlike most simple shift registers, the drivers' outputs are current sinks instead of current 
sources. This will be important when selecting the LEDs: In order to control each colour separately 
in an RGB LED, the cathodes have to be separated, while the anodes can be connected together.

\subsection{LEDs}

Like all the other parts, I ordered the LEDs from Digikey. Their website
contains an excellent tool, enabling the user to restrict the selection based on different criteria.
My requirements were the following:

\begin{itemize}

\item Colours: Red, green and blue
\item Current: 20 mA or more. Because the LEDs are within one pixel only $\frac{1}{100}$ of the
time, they are essentially pulse width modulated, reducing their effective brightness. In order to
still see the image clearly without darkening the room, they have to be bright enough.
\item Mounting type: Surface mount, so that I can route traces past the LEDs on the other side 
of the PCB
\item Size: At most 5 * 5 mm, preferably smaller, but still large enough to be able to solder them
by hand
\item Diffused lens, so the colours mix well even when viewing the LED directly instead of using
it for illumination.

\end{itemize}

These requirements reduced the category "LED Indication - Discrete" from 18940 to 130 products.
From those remaining products, I chose the one that was cheapest in a quantity of 160. This 
led me to an LED with the beautiful name CLVBA-FKA-CAEDH8BBB7A363.

\subsection{Capacitors}

A digital circuit like an LED driver can change its current consumption significantly within a
short amount of time. Because the connections to the power supply are relatively long, they have 
a substantial parasitic inductivity. So when the IC suddenly needs more current because all its
LEDs have just been switched on, this current is not available instantly. Worse yet, if the LEDs
are on and then are turned off, the current can't immediately stop flowing and will induce
dangerous voltage spikes.

In order to prevent those two situations from happening, I used capacitors. Namely, each LED 
driver has its own \SI{1}{\micro\farad}  ceramic capacitor.

\subsection{Motor}

In order for the whole device to rotate, I needed a motor. Luckily, I could salvage an appropriate
part from an old RC motor boat. It's a big and heavy brushed motor. The fact that it has a built-in
fan and was water-cooled in the boat leads me to think that its power output is much higher than 
what I need.

Because of its high torque, I could attach the motor directly to the rotating shaft instead of 
using gears or a belt drive. That way there are less moving parts and the motor can turn at a lower
rotational speed, both resulting in less wear and noise.

\subsection{Power Transmission}

To transmit power from the stationary power supply to the rotating part, I used slip rings. I made
a relatively simple version, consisting of two round pieces of copper sheet on the rotating end, 
and two long pieces of copper sheet on the stationary part.

The slip rings must be able to transmit the current necessary to light all LEDs up at once.
Assuming that each LED needs 20 mA, this current is about 9.6 A. Additionally, there is a small 
current draw by the microcontroller, but it will be negligible in comparison. Testing has shown 
that the slip rings don't heat up noticeably when conducting 10 A for several minutes.

In practice, however, this current is not reached since the vast majority of pixels are 
switched off when displaying a typical 3D image.

Because two pieces of copper sheet rubbing against each other aren't the most reliable electrical 
connection, I used a \SI{2200}{\micro\farad} electrolytic capacitor to smooth out the supply voltage. Even with it,
I found that the pressure between the two copper conductors had to be quite high for them to form
a good contact.

\newpage
\section{Circuit Diagram}

\begin{figure}[h]
\vspace{4mm}
\includegraphics[width=\textwidth]{./images/schematic.png}
\vspace{-8mm}
\caption{The finished circuit diagram}
\vspace{6mm}
\end{figure}

The circuit diagram --- as well as the PCB layout --- were created in Eagle. Its free version is
limited to two-layer boards of sizes up to \SI{10}{\centi\meter} by \SI{8}{\centi\meter}, but
that's more than enough for my purposes.

In figure 2, the final version of the schematic is pictured. For the sake of simplicity and 
readibility, the layout corresponds roughly to how the board will look.

I had to configure the footprints for the LED drivers and the LEDs by myself, which I did by
reading about the physical dimensions in the respective data sheet. For the resistors and 
capacitors, Eagle had pre-built footprints which I could use.

\subsection{Data Interface}

For transmitting data to the LED drivers, I used a simple row of 5 pins. Two of those pins are the
data lines: The input is connected to the first chip's input, and the output comes from the last
chip's output. The rest of the pins (clock, output enable and latch enable) are control lines,
which go to all three chips simultaneously. The order of these connections was determined when 
making the PCB layout.

\subsection{LED Drivers}

The LED drivers are arranged in a so-called daisy chain configuration, where the serial data output
of each chip is connected to the serial data input of the next chip. This is not only the case 
within one board, but also between the different boards: The data output of one board is connected
to the next board's input. That way, all of the 30 LED drivers can be controlled by a single SPI
bus without anything like a chip select signal. To send data to the chips, the microcontroller 
can simply send out 480 bit without interruption.

\def \iout {I\textsubscript{OUT}}
\def \rext {R\textsubscript{EXT}}

The external resistor which is connected to each LED driver sets the output
current. The following equation, taken from page 15 in the datasheet, describes the relation
between the output current (\iout) and the resistor's value (\rext):

% Lots of ugly hacks to have a non-italic font
\begin{ownmath}
\textrm{\iout} = 15 * \frac{\SI{1.25}{\volt}}{\si{\rext}}
\end{ownmath}

Let's find out the resistance necessary for the output current to be \SI{20}{\milli\ampere}:

\begin{ownmath}
\textrm{\rext} = 15 * \frac{\SI{1.25}{\volt}}{\SI{20}{\milli\ampere}} = \SI{937.5}{\ohm}
\end{ownmath}

The next smaller value within the E24 series is \SI{910}{\ohm}. By choosing this resistance, the 
output current is \SI{20.6}{\milli\ampere}, which is close enough to the desired 
\SI{20}{\milli\ampere}. I ordered 40 through-hole resistors of this value, because at the time I
had in mind to etch my own boards at home.

\subsection{LEDs}

The LEDs have their common anodes connected to \vdd. The cathodes are connected to a driver output
each, and are arranged red - green - blue (from left to right). Series resistors aren't needed 
because the LED drivers handle current regulation.

\section{PCB Design}

\begin{figure}[h]
\vspace{4mm}
\includegraphics[width=\textwidth]{./images/board.png}
\vspace{-4mm}
\caption{The PCB layout in Eagle. Red = upper copper layer, blue = lower copper layer,
green = vias, grey = silkscreen.} 
\vspace{6mm}
\end{figure}

\subsection{General Remarks}

The defining factor for the board layout was the pin layout of the LED drivers. The outputs are all
at the bottom of the chip, while the data interface is at the top. This led me to the layout I have
now: the LEDs are in a row at the bottom of the board where they can be directly connected to the driver 
outputs on the upper copper layer. The control lines are is routed on the back side of the board
behind the chips, with vias going to the front side to connect to each chip. The data lines go 
straight from one chip to the next one on the upper copper layer, with the last output going back
to the board's data interface on the back side. The resistors are placed on the right side of their
respective driver IC, jumping across the data traces.

The supply lines are placed beneath the row of LEDs. Had I placed the supply on the upper edge of the 
board, I still would've had to connect all the 16 anodes to \vdd. The driver chips, on the other hand, 
only need one connection to \vdd and GND each, so by placing the supply lines at the bottom of the board 
I have less traces running across the board vertically.

From a layout perspective, it would've made sense to put the GND line at the top of the board, since the
chips need one connection to that trace each, while the LEDs don't. However, I opted to put both supply
lines right above each other to reduce electromagnetic interference. If all the LEDs are switched on
suddenly and a lot of current starts to flow, a magnetic field builds up quickly around the supply traces.
This could induce bothering or even dangerous currents in the surrounding electronics. By having the two
supply lines close to each other, their magnetic fields largely cancel each other out. Another advantage
of having the supply at the bottom is that the distance to the delicate chips are a bit farther away.

\subsection{Space-saving measures}

In order for the production to be as cheap as possible as well as to minimise the rotating mass, I tried
to make the boards as small as possible. Ultimately, I managed to bring the size down to 
\SI{73}{\milli\meter} by \SI{21}{\milli\meter}

\end{document}

### 2016-05-13 paper.tex
%s/\\item/\t\\item/g
Finished chapter 2

@@
 Because two pieces of copper sheet rubbing against each other aren't the most reliable electrical 
-connection, I used a \SI{2200}{\micro\farad} electrolytic capacitor to smooth out the supply voltage. Even with it,
-I found that the pressure between the two copper conductors had to be quite high for them to form
-a good contact.
+connection, I used a \SI{2200}{\micro\farad} electrolytic capacitor to smooth out the supply 
+voltage. Even with it, I found that the pressure between the two copper conductors had to be quite
+high for them to form a good contact.
 
@@
 at the bottom of the chip, while the data interface is at the top. This led me to the layout I have
-now: the LEDs are in a row at the bottom of the board where they can be directly connected to the driver 
-outputs on the upper copper layer. The control lines are is routed on the back side of the board
-behind the chips, with vias going to the front side to connect to each chip. The data lines go 
-straight from one chip to the next one on the upper copper layer, with the last output going back
-to the board's data interface on the back side. The resistors are placed on the right side of their
-respective driver IC, jumping across the data traces.
+now: the LEDs are in a row at the bottom of the board where they can be directly connected to the 
+driver outputs on the upper copper layer. The control lines are is routed on the back side of the 
+board behind the chips, with vias going to the front side to connect to each chip. The data lines 
+go straight from one chip to the next one on the upper copper layer, with the last output going 
+back to the board's data interface on the back side. The resistors are placed on the right side of 
+their respective driver IC, jumping across the data traces.
 
-The supply lines are placed beneath the row of LEDs. Had I placed the supply on the upper edge of the 
-board, I still would've had to connect all the 16 anodes to \vdd. The driver chips, on the other hand, 
-only need one connection to \vdd and GND each, so by placing the supply lines at the bottom of the board 
-I have less traces running across the board vertically.
+The supply lines are placed beneath the row of LEDs. Had I placed the supply on the upper edge of 
+the board, I still would've had to connect all the 16 anodes to \vdd. The driver chips, on the 
+other hand, only need one connection to \vdd and GND each, so by placing the supply lines at the 
+bottom of the board I have less traces running across the board vertically.
 
-From a layout perspective, it would've made sense to put the GND line at the top of the board, since the
-chips need one connection to that trace each, while the LEDs don't. However, I opted to put both supply
-lines right above each other to reduce electromagnetic interference. If all the LEDs are switched on
-suddenly and a lot of current starts to flow, a magnetic field builds up quickly around the supply traces.
-This could induce bothering or even dangerous currents in the surrounding electronics. By having the two
-supply lines close to each other, their magnetic fields largely cancel each other out. Another advantage
-of having the supply at the bottom is that the distance to the delicate chips are a bit farther away.
+From a layout perspective, it would've made sense to put the GND line at the top of the board, 
+since the chips need one connection to that trace each, while the LEDs don't. However, I opted to 
+put both supply lines right above each other to reduce electromagnetic interference. If all the 
+LEDs are switched on suddenly and a lot of current starts to flow, a magnetic field builds up 
+quickly around the supply traces. This could induce bothering or even dangerous currents in the 
+surrounding electronics. By having the two supply lines close to each other, their magnetic fields 
+largely cancel each other out. Another advantage of having the supply at the bottom is that the 
+distance to the delicate chips are a bit farther away.
 
@@
 
-In order for the production to be as cheap as possible as well as to minimise the rotating mass, I tried
-to make the boards as small as possible. Ultimately, I managed to bring the size down to 
-\SI{73}{\milli\meter} by \SI{21}{\milli\meter}
+In order for the production to be as cheap as possible as well as to minimise the rotating mass, I 
+tried to make the boards as small as possible. Ultimately, I managed to bring the size down to 
+\SI{73}{\milli\meter} by \SI{21}{\milli\meter}.
+
+In order to achieve this, but also to avoid large gaps between the pixels, I placed the LEDs as 
+close as possible to each other. The datasheet claims a width of \SI{2.8 +- 0.2}{\milli\meter}, 
+so I decided to place the LEDs at \SI{3}{\milli\metre} increments. On a grid of \SI{12.5}{mil}, 
+the next larger distance is \SI{125}{mil}, which is equal to \SI{3.175}{mm}.
+
+The connections to the cathodes of the red LED needs to pass by the other two cathodes, since 
+they're located on the lower side of the LED package. Because the LEDs are so close to each other, 
+the only option was to use the lower copper layer. The only place to go back to the upper copper 
+layer was right beneath the driver chips.
+
+Initially, I had placed the capacitors right above the ICs. However, I could save a small bit of
+vertical board size by placing them at an offset to the right. That way I could push them down 
+until where the pins begin, which is a bit further down than the edge of the chip itself.
+
+\subsection{Thermal Considerations}
+
+A positive side effect of placing some vias beneath the chip is their thermal conductivity. That
+way, the chips' waste heat can be dissipated more effectively. The maximum power output of the 
+chips --- when all the attached LEDs are turned on --- is as follows:
+
+\begin{ownmath}
+\textrm{P\textsubscript{MAX}} = (\SI{5}{\volt} - \textrm{V\textsubscript{F}}) * 
+\SI{20.6}{\milli\ampere\per output} * \SI{16}{outputs} = \SI{867.9}{\milli\watt}
+\end{ownmath}
+
+V\textsubscript{F} is different depending on the colour of the LED, so I just used the average
+value, which is \SI{2.633}{\volt}. The actual values are \SI{1.9}{\volt} for red and \SI{3}{\volt}
+for green and blue.
+
+A power output of almost a watt is not to be underestimated, and tests have shown that that the
+boards do get quite hot when operated at full power. In reality, however, the situation is much
+less severe because the LEDs are typically turned on only during a fraction of the rotation.
+Furthermore, the quick rotation results in excellent air circulation, which aids the dissipation
+of heat.
 

### 2016-05-14 paper.tex
Added & finished chapter 3

@@
 \vspace{4mm}
-\includegraphics[width=\textwidth]{./images/board.png}
+\includegraphics[width=\textwidth]{./images/board-layout.png}
 \vspace{-4mm}
@@
 
+\chapter{Fabrication}
+
+\section{Circuit Boards}
+
+\subsection{Fabrication}
+
+There's no shortage of firms who produce PCBs according to a custom layout. One of the cheapest 
+providers to ship small production runs within a reasonable timeframe and with good production
+quality turned out to be OSHPark. They only produce in multiples of 3, so I ordered 12 instead of
+10 (which I'd have done anyway to have some spare boards). I submitted my order on the 5th of 
+October 2014 and paid US\$ 47. Ten days later I received a message that the boards were finished,
+and after another ten days of waiting they arrived at my doorstep.
+
+\begin{figure}[h]
+\vspace{3mm}
+\includegraphics[width=\textwidth]{./images/board-finished.jpeg}
+\vspace{-8mm}
+\caption{A finished circuit board} 
+\end{figure}
+
+\subsection{Assembly}
+
+As soon as I got the boards in the mail, I started soldering the parts to them. I started with the
+LEDs to avoid having to solder in the tight gap that would later be formed between the LEDs and
+their driver chips. It took me two hours to solder the first board, but with increasing practice
+the time for one row of LEDs got down to 25 or 30 minutes.
+
+The chips were a lot easier to solder than the LEDs, despite the distance between the individual
+pins only being \SI{0.635}{\milli\meter}. I could simply drown all the pins in solder, and then 
+wick away all the solder bridges. That way I could solder twelve pins within the blink of an eye.
+
+\subsection {Testing and Debugging}
+
+With the boards finished, I wanted to know whether or not they work. In order to find out, I wrote
+a little program on the microcontroller to light the LEDs up in all possible colours consecutively.
+It did not work at the first try, so I used an oscilloscope to see if data was being sent at all. 
+This was the case, the SPI library on the Teensy performed its task flawlessly.
+
+It turned out that the culprit were the bad contacts. Instead of soldering the wires onto the board
+and the microcontroller, I simply inserted them into the respective holes. This caused the 
+connection to have too much parasitic capacitance, and the signals (which were being sent at 4 MHz)
+did not have time to propagate up to their destination.
+
+I soldered the cables on properly and sure enough, the board worked as intended. I repeated this
+procedure for all ten boards to see if they all worked as well as the first one, which wasn't the 
+case.
+
+Because of reasons unbeknownst to me, one particular LED driver started turning on all of its LEDs
+at a moderate brightness as soon as I powered it up --- even though no data was being sent at all.
+At first I thought that the serial data input had a solder bridge pulling it up to \vdd, and
+immediately sending a whole lot of 1's upon powering up the board. However, the data input pin was
+not close enough to a positive supply line for this to be an issue. Additionally, there would have 
+to be a clock signal as well as a pulse on the latch pin for the LEDs to light up. I gave up and 
+assumed that the chip was faulty, and when I replaced it with a spare part everything worked 
+perfectly.
+
+Apart from that, there were no major issues. I had a few LEDs which didn't light up when they 
+should have. One of them had a solder bridge between the common anode and the cathode of the red
+LED, which made a voltage drop across the red LED impossible. Two were simply soldered on poorly
+and weren't connected to the drivers at all. A few other LEDs didn't work and I found no obvious
+issues. I replaced them with spares, which did the trick.
+
+\section{Physical Assembly}
+
+\subsection{Frame}
+
+The frame was built out of beech wood strips with a square cross section and a side length of 
+\SI{1.5}{\centi\meter}. The baseplate is made out of plywood and has the dimensions 
+\SI{15}{\centi\meter} by \SI{15}{\centi\meter} by \SI{1}{\centi\meter}. On it, I mounted two 
+vertical wood strips with a length of \SI{20}{\centi\meter}, which were both stabilise by three
+struts. A horizontal board, mounted to the vertical strips at a height of \SI{20}{\centi\meter}, 
+holds the motor.
+
+On the top of the two vertical strips I placed another strip to hold the shaft in its position. 
+A simple hole served as a bearing, which I later replaced with a plastic insert (kindly sponsored
+by IGUS).
+
+\subsection{Slip Rings}
+
+In order to supply the rotating parts with electricity, I mounted a wooden disc right at the bottom
+of the shaft. On both sides of that disc I glued on a round piece of copper sheet, which I then
+connected to cables going up to where the electronics would later be.
+
+On the stationary end, two long pieces of copper sheet serve as contacts. Because of their 
+spring-like properties, they exert a certain amount of pressure on the rotating disc.
+
+Despite this, a considerable amount of pressure was necessary to form a good contact. The contact
+pressure proved to be a primitive, but usable way of roughly adjusting the rotational speed of
+the display. Sadly it's quite an inefficient approach: The motor uses a lot of current when 
+friction is applied to its load. It becomes noticeably warm after a few minutes of use, and 
+measurements on the current limiting resistors have shown that it draws about \SI{6.5}{\ampere} at
+\SI{5}{\volt}.
+
+This constant pressure, in addition to the fact that copper is quite a soft metal, results in 
+considerable wear on the slip rings. Luckily, the stationary ends can be replaced without much 
+hassle. The rotating part would be a pain to replace, but since wear is distributed across the 
+whole copper disc, it will be a long time before this needs to be done.
+
+\subsection{Controller}
+
+The microcontroller is mounted on another wooden disc, located a few centimetres above the slip
+rings. It also hosts the hall sensor, which is used to measure the rotational speed and calculate
+the position of the rotating part at any given moment. The corresponding magnet is mounted on a 
+piece of threaded rod, which itself is mounted on the same board as the motor.
+
+Additionally, the \SI{2200}{\micro\farad} capacitor is located on the upper wooden disc. In the 
+end, this place has become a bit of a mess, especially with all the cables going up to the circuit 
+boards. Luckily this isn't visible while the apparatus is running.
+
+\subsection{Circuit Boards}
+
+The boards were already equipped with components, but what was missing was the hold to mount them
+on the shaft. The diameter of this shaft is \SI{4}{\milli\meter}. In order to find a good hole 
+size, I drilled a few test holes on a spare board. A hole of \SI{4}{\milli\meter} was so small for
+the boards to fit on the shaft, while \SI{4.5}{\milli\meter} was obviously too big. After 
+chamfering the top end of the shaft, the board with a \SI{4}{\milli\meter} hole just barely could
+slide on the shaft. This tight fit proved to be advantageous, because torque can now be transmitted
+reliably from the shaft to the boards, and they don't slide around.
+
+\chapter{Programming}
+
 \end{document}

### 2016-05-14 paper.tex
Removed trailing whitespace
Nicer signature field
Fixed math error

@@
 \vspace{10mm}
-\noindent Location and Date: \hspace{0.35\textwidth} Signed:\\
-\rule[3mm]{\textwidth}{0.5pt}
+\rule{\textwidth}{0.5pt}
+\begin{minipage}[t]{0.5\textwidth}
+Location and Date
+\end{minipage}
+\begin{minipage}[t]{0.5\textwidth}
+\hspace{-1mm}Signature
+\end{minipage}
 
@@
 \textrm{P\textsubscript{MAX}} = (\SI{5}{\volt} - \textrm{V\textsubscript{F}}) *
-\SI{20.6}{\milli\ampere\per output} * \SI{16}{outputs} = \SI{867.9}{\milli\watt}
+\SI{20.6}{\milli\ampere\per output} * \SI{16}{outputs} = \SI{780}{\milli\watt}
 \end{ownmath}
@@
 
-A power output of almost a watt is not to be underestimated, and tests have shown that that the
-boards do get quite hot when operated at full power. In reality, however, the situation is much
-less severe because the LEDs are typically turned on only during a fraction of the rotation.
-Furthermore, the quick rotation results in excellent air circulation, which aids the dissipation
-of heat.
+A power output of \SI{780}{\milli\watt} is not to be underestimated on such a small PCB, and tests
+have shown that that the boards do get quite hot when operated at full power. In reality, however,
+the situation is much less severe because the LEDs are typically turned on only during a fraction
+of the rotation. Furthermore, the quick rotation results in excellent air circulation, which aids
+the dissipation of heat.
 

### 2016-05-15 paper.tex
wording
Finished programming chapter

@@
 
+% Display SI units
 \usepackage{siunitx}
@@
 
-Because of that, I ditched the idea of 8-bit colour depth and instead went for the simpler approach
+Because of these problems, I ditched the idea of 8-bit colour depth and instead went for the simpler approach
 of one-bit colour depth. Both problems are solved by switching to 1 bit: The data rate doesn't have
@@
 \includegraphics[width=\textwidth]{./images/board-layout.png}
-\vspace{-4mm}
+\vspace{-6mm}
 \caption{The PCB layout in Eagle. Red = upper copper layer, blue = lower copper layer,
@@
 surrounding electronics. By having the two supply lines close to each other, their magnetic fields
-largely cancel each other out. Another advantage of having the supply at the bottom is that the
-distance to the delicate chips are a bit farther away.
+largely cancel each other out.
 
@@
 
-Initially, I had placed the capacitors right above the ICs. However, I could save a small bit of
-vertical board size by placing them at an offset to the right. That way I could push them down
-until where the pins begin, which is a bit further down than the edge of the chip itself.
+Initially, I had placed the decoupling capacitors right above the ICs. However, I could save a 
+small bit of vertical board size by placing them at an offset to the right. That way I could push 
+them down until where the pins begin, which is a bit further down than the edge of the chip itself.
 
@@
 
+There were two factors I had to carefully consider while arranging the boards:
+
+\begin{itemize}
+\item The boards had to be balanced out to minimise static and dynamic unbalance
+\item The amount of boards obstructing other boards had to be minimal
+\end{itemize}
+
+In the end, I decided to arrange the boards in a double-spiral arrangement. This was the only
+arrangement I could think of that (theoretically) has no static imbalance at all. Because each
+board has another board on the exact opposite side of the shaft, there are five pairs of boards
+where one board statically balances out the other. However, each pair has some dynamic inbalance 
+because the two boards aren't mounted at the same height. This could've been minimised by 
+alternating the order of boards between each pair. However, this wasn't necessary because the 
+vibrations were within reasonable bounds.
+
+The visibility is not too bad either: On the viewer-facing half of the display, all LEDs are 
+constantly visible. On the backside, it is possible that one group of boards is right before the
+other one, causing some layers of the image to disappear.
+
 \chapter{Programming}
 
+\section{Used Software}
+
+The following program has been compiled with the Arduino IDE 1.0.6 and sent to the Teensy by the
+Teensyduino 1.20 plugin. However, the most part of the program has been written in Sublime Text 2
+with the stino plugin. This combination has a few advantages over simply using the Arduino IDE to
+write all code, namely there's automatic code completion, support for several cursors, and a 
+search-and-replace function with regex suport. I installed and used these programs on Mac OS 10.9
+and later 10.10.
+
+\section{Timing}
+
+\subsection{Theory}
+
+In order for the image to stay still despite the rotational speed not being completely constant, 
+the frequency of LED refreshed must be constantly adjusted based on the rotational period. Because
+there are 100 pixels per revolution, the time to display one pixel must be $\frac{1}{100}$ of the
+rotational period.
+
+This value can be measured easily by using the hall sensor. By dividing this value by 100, we can 
+obtain the duration of one pixel for the current speed. If the rotational speed changes, that's a
+bit of a problem: The measured value is from the previous rotation, but in the current rotation, 
+the device already turns slightly faster. This causes the image to be slightly warped right after
+powering up the display. However, once it has reached temrminal velocity, this effect is no longer
+noticeable, because now the subsequent rotations have about the same period.
+
+\subsection{Implementation}
+
+This behaviour is implemented with an interrupt that runs each time the magnet is detected (i.e.
+after each rotation). Within that interrupt, several things happen:
+
+First, we read a variable called \texttt{sinceMagnet}. As its name says, it keeps track of 
+the number of microseconds since the last time the magnet was detected. It's of the special type
+\texttt{elapsedMicros}, which means that it's automatically incremented each microsecond. This
+value is divided by 100 and stored in \texttt{microsPerPixel}, where it represents the amount of
+time that any given pixel should be displayed for.
+
+This variable is now reset to 0 to start the new rotation. The same is done to another
+variable, \texttt{nextPixelMicros}. It contains the next point in time when the pixels
+should be refreshed, with respect to the start of the rotation (which just happened).
+
+In the main loop of the program, this variable is continuously compared to \texttt{sinceMagnet}. If
+sinceMagnet is bigger, the LEDs are refreshed by calling \texttt{sendData}. Within that function,
+\texttt{nextPixelMicros} is incremented by \texttt{microsPerPixel}, causing the program to wait for
+the display to turn another $\frac{1}{100}$ of a full circle. After that, it'll display the next
+pixel.
+
+The implementation of the described behaviour can be found in the appendix, where the whole source
+code is to be found. Specifically, what I've just talked about is contained in the
+\texttt{timerUpdate} function, as well as in the main \texttt{loop}.
+
+\section{Data storage and display}
+
+The previously mentioned function \texttt{sendData} needs to be able to know which bytes to send
+based on one parameter, which is the current angle of of the display. In this section, I will
+explain how I achieved this.
+
+To save the image on the microcontroller, I used a simple three-dimensional array containing 100 *
+10 * 6 bytes of data. The first index stands for the angle of a pixel in hundredths of a circle,
+the second index corresponds to the height and the third one describes the position of each byte 
+on a row of LEDs. Because there are 48 LEDs per row, 6 bytes can be used to describe the row's
+state without wasting any storage space. The only way to be more efficient in terms of storage
+would be to compresss data, which would require significantly more computing power.
+
+Because the PCBs are not right above each other, the data that is being sent out in one refresh
+don't correspond to the pixels that are above each other in the image. In order avoid having to
+account for the angular offsets of each board while creating an image, I made the program do that
+work automatically.
+
+Each pair of two boards are mounted on the exact opposite side, making the angle between them
+\SI{180}{\degree} or 50 pixels. Each subsequent pair is additionally rotated by \SI{28.8}{\degree}
+which is equivalent to 8 pixels.
+
+The program accounts for this arrangement by using a simple for loop. It iterates over five pairs
+of boards. At its start, the program calculates the angle and height of the first board based on
+the control variable. After the data has been sent, it then moves to the second board by adding 50
+to the angle and incrementing the board counter, and feeds the second board with data too.
+
+After these two additions, the angle value will go over 99 in many cases, which is the highest
+possible array index. Since the pixel at position $100+n$ is the same as the one at position $n$, I
+can pervent causing undefined behaviour by applying a modulus 100 operator to the angle value
+before using it to access the array.
+
+\section{Miscellaneous}
+
+While the program works the way it is described above, there are a few small things that I managed
+to further improve.
+
+Because I selected the position of the boards with respect to the hall sensor arbitrarily, the
+image was turned around. To fix that, I added a constant value to all angle values, so now the
+first pixel is displayed exactly at 0° instead of whereever the hall sensor detects the magnet.
+
+Whenever I started up the program, it started displaying the first pixel. It even did so when I
+connected the Teensy to my laptop via USB. This was dangerous because neither the Teensy nor the
+laptop's USB port can handle a lot of current, and if more than a handful of LEDs were switched on,
+the Teensy could've been damaged.
+
+To fix this I added a boolean variable called \texttt{running}, which is set to \texttt{false} by
+default and only flipped when the magnet is first detected. That way the program only starts
+running if the display is actually rotating, at which point we can be sure that power is supplied
+through the sliprings rather than the USB port.
+
+The images displayed were constantly jittering around instead of staying still like they should. It
+took me a long time to find the source of this bug, but I managed to fix it in time for the swiss
+national competition organised by "Swiss Youth in Science".
+
+The problem was that sometimes, the magnet was detected while \texttt{sendData} was still running.
+Because the hall sensor's interrupt had a higher priority than the timer interrupt I had used to
+send data, control was immediately given to the former. The \texttt{timerUpdate} function was
+executed normally, resetting \texttt{currentPixel} to zero and reading out the duration of the last
+rotation. However, after that, control was given bach to the \texttt{sendData} function. At the end
+of that function, \texttt{currentPixel} was incremented by one. That way, the \texttt{currentPixel}
+variable already had a value of one when it should've stayed at 0. For the entirety of the next
+rotation, \texttt{currentPixel} was off by one, resulting in the image quickly turning by
+\SI{3.6}{\degree}.
+
 \end{document}

### 2016-05-16 paper.tex
Unindented \itemize, removed trailing \s
Another chapter, nicer math, etc

@@
 \usepackage{xspace}
-\def \iZc {I\textsuperscript{2}C\xspace}
+\def \itwoc {I\textsuperscript{2}C\xspace}
 \def \vdd {V\textsubscript{DD}}
@@
 % Display SI units
-\usepackage{siunitx}
+\usepackage[separate-uncertainty=true, multi-part-units=single]{siunitx}
 
@@
 Neither of these is being produced on a grand scale, instead they are --- like most 2D versions
---- side projects of makers or engineers.
+--- side projects of hobbyists or engineers.
 
@@
 
-	\item Resolution of 16 * 100 * 10 pixels (radius * circumference * height)
+\item Resolution of $16 \cdot 100 \cdot 10$ pixels (radius $\cdot$ circumference $\cdot$ height)
 \item Each pixel consists of an RGB LED
@@
 
-	\item Controlled using \iZc  instead of SPI
+\item Controlled using \itwoc  instead of SPI
 
-Because \iZc uses pullup resistors instead of a push-pull drive, it takes a little moment for the
-logic level to return to its normal state. In order to ensure correct data transmission, it is
-necessary to set a relatively slow clock speed. The two \iZc interfaces of the Teensy only support
-clock speeds of 2.4 MHz --- and because \iZc requires some overhead data to send out the slave's
-addresses, the actual data rate will be even lower.
+Because \itwoc uses pullup resistors instead of a push-pull drive, it takes a little moment for
+the logic level to return to its normal state. In order to ensure correct data transmission, it is
+necessary to set a relatively slow clock speed. The two \itwoc interfaces of the Teensy only
+support clock speeds of up to 2.4 MHz --- and because \itwoc requires some overhead data to send
+out the slave's addresses, the actual data rate will be even lower.
 
@@
 % Yes it's a mess but \si makes it a unit, so it is not in cursive while still being small enough
-24\,\frac{\si{bit}}{\si{pixel}} * 16000\,\frac{\si{pixels}}{\si{rotation}} * 30\,\frac{\si{rotations}}{\si{second}} = \SI{11.52}{Mbps}
+24\,\frac{\si{bit}}{\si{pixel}} \cdot 16000\,\frac{\si{pixels}}{\si{rotation}} \cdot
+30\,\frac{\si{rotations}}{\si{second}} = \SI{11.52}{Mbps} 
 \end{ownmath}
 
-This means that I'd need at least 5 parallel \iZc lines to control all the LEDs with 8 bit colour
+This means that I'd need at least 5 parallel \itwoc lines to control all the LEDs with 8 bit colour
 depth. Because each Teensy only has two of them, I'd need three Teensy boards, which would be
@@
 
-Because of these problems, I ditched the idea of 8-bit colour depth and instead went for the simpler approach
-of one-bit colour depth. Both problems are solved by switching to 1 bit: The data rate doesn't have
-to be as high (and there are lots of chips that are controlled by SPI), and PWM is not necessary
-anymore.
+Because of these problems, I ditched the idea of 8-bit colour depth and instead went for the
+simpler approach of one-bit colour depth. Both problems are solved by switching to 1 bit: The data
+rate doesn't have to be as high (and there are lots of chips that are controlled by SPI), and PWM
+is not necessary anymore.
 
@@
 
-After quite some searching around, I decided to use the TLC5927 by Texas Instruments. This is
-essentially a shift register with some extra features. Like every shift register, it has a serial
-data input (SDI), a serial data output (SDO), and a clock pin (CLK). Additionally, there is a latch
-pin (LE), which serves the purpose of only refreshing the state of the LEDs once all data has been
-sent.
+After quite some searching around, I decided to use the TLC5927 by Texas Instruments. It is
+essentially a shift register with some extra features. Like every normal shift register, it has a
+serial data input (SDI), a serial data output (SDO), and a clock pin (CLK). Additionally, there is
+a latch pin (LE), which serves the purpose of only refreshing the state of the LEDs once all data
+has been sent.
 
@@
 
-Like all the other parts, I ordered the LEDs from Digikey. Their website
-contains an excellent tool, enabling the user to restrict the selection based on different criteria.
-My requirements were the following:
+Like all the other parts, I ordered the LEDs from Digikey. Their website contains an excellent
+tool, enabling the user to restrict the selection based on different criteria. My requirements
+were the following:
 
@@
 \item Colours: Red, green and blue
+
 \item Current: 20 mA or more. Because the LEDs are within one pixel only $\frac{1}{100}$ of the
@@
 still see the image clearly without darkening the room, they have to be bright enough.
-	\item Mounting type: Surface mount, so that I can route traces past the LEDs on the other side
-of the PCB
-	\item Size: At most 5 * 5 mm, preferably smaller, but still large enough to be able to solder them
-by hand
+
+\item Mounting type: Surface mount, so I can still use the other side of the board
+
+\item Size: At most $\SI{5}{\milli\meter}$ by $\SI{5}{\milli\meter}$, preferably smaller, but
+still large enough to solder by hand
+
 \item Diffused lens, so the colours mix well even when viewing the LED directly instead of using
@@
 
-These requirements reduced the category "LED Indication - Discrete" from 18940 to 130 products.
+These requirements reduced the category "LED Indication - Discrete" from 18940 to 130 items.
 From those remaining products, I chose the one that was cheapest in a quantity of 160. This
@@
 \begin{ownmath}
-\textrm{\iout} = 15 * \frac{\SI{1.25}{\volt}}{\si{\rext}}
+\textrm{\iout} = 15 \cdot \frac{\SI{1.25}{\volt}}{\si{\rext}}
 \end{ownmath}
@@
 \begin{ownmath}
-\textrm{\rext} = 15 * \frac{\SI{1.25}{\volt}}{\SI{20}{\milli\ampere}} = \SI{937.5}{\ohm}
+\textrm{\rext} = 15 \cdot \frac{\SI{1.25}{\volt}}{\SI{20}{\milli\ampere}} = \SI{937.5}{\ohm}
 \end{ownmath}
@@
 In order to achieve this, but also to avoid large gaps between the pixels, I placed the LEDs as
-close as possible to each other. The datasheet claims a width of \SI{2.8 +- 0.2}{\milli\meter},
+close as possible to each other. The datasheet claims a width of \SI{2.8 \pm 0.2}{\milli\meter},
 so I decided to place the LEDs at \SI{3}{\milli\metre} increments. On a grid of \SI{12.5}{mil},
@@
 \begin{ownmath}
-\textrm{P\textsubscript{MAX}} = (\SI{5}{\volt} - \textrm{V\textsubscript{F}}) *
-\SI{20.6}{\milli\ampere\per output} * \SI{16}{outputs} = \SI{780}{\milli\watt}
+\textrm{P\textsubscript{MAX}} = (\SI{5}{\volt} - \textrm{V\textsubscript{F}}) \cdot
+\SI{20.6}{\milli\ampere\per output} \cdot \SI{16}{outputs} = \SI{780}{\milli\watt}
 \end{ownmath}
@@
 
-To save the image on the microcontroller, I used a simple three-dimensional array containing 100 *
-10 * 6 bytes of data. The first index stands for the angle of a pixel in hundredths of a circle,
-the second index corresponds to the height and the third one describes the position of each byte 
-on a row of LEDs. Because there are 48 LEDs per row, 6 bytes can be used to describe the row's
-state without wasting any storage space. The only way to be more efficient in terms of storage
-would be to compresss data, which would require significantly more computing power.
+To save the image on the microcontroller, I used a simple three-dimensional array containing $100
+\cdot 10 \cdot 6$ bytes of data. The first index stands for the angle of a pixel in hundredths of
+a circle, the second index corresponds to the height and the third one describes the position of
+each byte on a row of LEDs. Because there are 48 LEDs per row, 6 bytes can be used to describe the
+row's state without wasting any storage space. The only way to be more efficient in terms of
+storage would be to compresss data, which would require significantly more computing power.
 
@@
 
+\chapter{Operation}
+
+\section{Power Supply}
+
+Initially I had planned to supply the PCBs with \SI{5}{\volt}. However, the Teensy sends its
+signals at \SI{3.3}{\volt}. The driver chips, on the other hand, only read a logical high when a
+signal is over $0.7 \cdot \textrm{\vdd}$. At \SI{5}{\volt}, this value would be \SI{3.5}{\volt}.
+So I tried powering the boards with only \SI{3.3}{\volt}, and everything worked perfectly. Better
+yet, the boards weren't getting as hot anymore.
+
+According to the manufacturer, the Teensy needs at least \SI{3.7}{\volt} to operate, presumably
+because it uses a linear regulator to get a stable voltage ov \SI{3.3}{\volt}. However, there were
+no problems when I started using \SI{3.3}{\volt}. Even though the regulator probably reduces that
+voltage to less than \SI{3}{\volt}, the processor still runs at its full clock speed of
+\SI{96}{\mega\hertz}.
+
+Unfortunately, the PC power supply which I used for the project broke down. As I later found out,
+the culprit was the motor consuming too much current. At that time, I didn't use any series
+resistance to limit the current flowing through the motor, which I powered with \SI{3.3}{\volt}.
+This has caused the inrush current to be greater than what the power supply could handle.
+
+To remedy this, I used a pair of long, thin wires to connect to the motor to provide some
+resistance. I later replaced these with two \SI{150}{\milli\ohm} power resistors. Now there are two
+ways to power the motor: I can use \SI{5}{\volt} and connect the resistors in series for a total
+resistance of \SI{300}{\milli\ohm}, or I can power it with \SI{3.3}{\volt} and connect the
+resistors in parallel, resulting in a resistance of \SI{75}{\milli\ohm}. The latter method is more
+efficient, but it causes the acceleration at startup to be slower.
+
+\section{Mechanical Problems}
+
+Because the slip rings weren't fixed in their position, they could slide out of their slots after
+a few minutes of operation. I fixed this issue by mounting a small wooden board onto these slots
+(perventing lateral movement), and routing the cables going from the power supply to the sliprings
+through two holes in that board (perventing lengthwise movement).
+
+The boards weren't exactly aligned at the \SI{28.8}{\degree} that I specified. This caused the
+image to be displayed crookedly: Pixels that should've been right above each other were shifted
+sideways by a small distance. Whenever I adjusted the boards, they would be misaligned again after
+using the display for some time. To fix this, I aligned them correctly one last time and used
+enough solder on the power cables running through the boards for them to form a solid connection.
+
+The bolt nut that held the motor in place started coming loose because of the vibrations. To
+pervent that from happening again, I used a second nut and tightened it against the first one.
+
 \end{document}

### 2016-05-17 paper.tex
Conclusions & outlook

@@
 
-\item Size: At most $\SI{5}{\milli\meter}$ by $\SI{5}{\milli\meter}$, preferably smaller, but
-still large enough to solder by hand
+\item Size: At most \SI{5}{\milli\meter} by \SI{5}{\milli\meter}, preferably smaller, but still
+large enough to solder by hand
 
@@
 
-A positive side effect of placing some vias beneath the chip is their thermal conductivity. That
-way, the chips' waste heat can be dissipated more effectively. The maximum power output of the
-chips --- when all the attached LEDs are turned on --- is as follows:
+A positive side effect of placing some vias beneath the chip is their thermal conductivity, which
+helps disspiate the chip's excess heat. The maximum power output of the chips --- when all the
+attached LEDs are turned on --- is as follows:
 
@@
 running if the display is actually rotating, at which point we can be sure that power is supplied
-through the sliprings rather than the USB port.
+through the slip rings rather than the USB port.
 
@@
 a few minutes of operation. I fixed this issue by mounting a small wooden board onto these slots
-(perventing lateral movement), and routing the cables going from the power supply to the sliprings
+(perventing lateral movement), and routing the cables going from the power supply to the slip rings
 through two holes in that board (perventing lengthwise movement).
@@
 
+\chapter{Conclusions}
+
+\section{Possible Improvements}
+
+\subsection{Controller and Data Transmission}
+
+The Teensy 3.1 is currently mounted quite loosely on its wooden disc. A piece of wire goes through
+two unoccupied contacts as well as through two holes in the disc. In addition, the cables forming
+the SPI bus aren't laid out neatly at all; many are too long and don't take the direct route.
+
+This situation could be improved by replacing the upper wooden disc with another printed circuit
+board. All connections could be routed elegantly on the PCB, and the SPI interfaces could be
+formed by two ribbon cables. The hall sensor could be properly soldered on instead of fixing it in
+place using hot glue.
+
+Between the individual LED boards, ribbon cables could also be considered. However, the only
+connections that always go to the board right above are CLK, LE and OE, while the serial data
+inputs and outputs have to be connected to the board on the opposite side. These connections would
+still have to be formed by individual cables.
+
+\subsection{Slip Rings}
+
+Because the slip rings are quite rough and not completely horizontal, a lot of pressure is needed
+to form a consistent contact. This leads to high friction between the two contacts, which then
+results in quick wear as well as a lot of friction.
+
+The main reason why the display is not energy efficient at all are the slip rings. At the start of
+the project, I had the idea of using a \SI{3.5}{\milli\meter} phone connector to transmit power.
+However, I abolished that plan because of two reasons. Firstly, I had concerns that such a small
+connector would have trouble conducting the maximum current of \SI{10}{\ampere}. This would most
+likely not have been a problem, since only a small part of all pixels are turned on in a typical
+image, which causes the average current to be a small fraction of the maximum value.
+
+The second problem would've been the difficulty of mounting a phone connector on the display. I
+would've had to place it at the end of the shaft, where it would've also served as a bearing. In
+order to minimise the forces acting upon the plug, it would've had to be mounted exactly in the
+center of the shaft. I couldn't come up with a solid way to connect both of the phone plugs' ends
+to the display.
+
+A much more elegant solution would've been to use induction to transfer power. By having a
+stationary and a rotating coil close together, I could've passed a high-frequency alternating
+current through the stationary one, causing current to flow in the rotating one as well. The
+advantages of such an approach would be substantial: The bearings would be the only contact point
+between the rotating and stationary part, causing friction to be minimal. I could've also
+completely eliminated the annoying, broadband noise currently emitted by the slip rings.
+
+I didn't implement wireless power transfer because I felt like the extra work wasn't worth it,
+especially considering this was a prototype. If I wanted to produce this display en masse,
+inductive power transfer would definitely be the way to go as it's the only way to make the
+display somewhat quiet.
+
+\subsection{Motor}
+
+I decided to drive the display directly without any kind of gear reduction because that was the
+simplest solution. I only could do so because the motor was strong enough. Unfortunately, it
+consumes a substantial amount of current.
+
+If I wanted to sell my display commercially, I'd use a smaller motor, drive it at a higher
+rotational speed and gear it down. To save energy, I would also implement one of the above
+imporvements concerning power transmission.
+
+\section{Future Outlook}
+
+\subsection{Animated images}
+
+Something I had in mind since the beginning of this project were animated images. The Teensy 3.1
+has \SI{256}{\kilo\byte} of memory, which would fit 42 images of \SI{6}{\kilo\byte} each. This
+corresponts to roughly a second of playback. In order to display longer animations, there are
+several workarounds: Instead of showing one image that spans the entirety of the rotation, I can
+instead only save an image that is 25 pixels and display it four times.
+
+In order to get around the limit of flash memory, external storage would be necessary. Even with
+only \SI{1}{\giga\byte}, I could save 166666 frames, thus removing any practical limit in terms of
+playback time.
+
+\subsection{Interactive Programs}
+
+It would be interesting to not only display predefined images, but to actually run a program that
+can respond to user input, like a video game. In order to communicate with the external world, a
+wireless interface would be necessary. Readily available solutions include small WLAN or bluetooth
+modules, or the transmitters and receivers commonly used in radio-controlled aircraft.
+
+\subsection{Applications}
+
+Because of the low resolution and colour depth, the high level of noise, quick wear and the
+inability to send images to the display while running, this display is not suited as a replacement
+for your everyday computer or TV screen. However, these problems can all be fixed completely or
+partially, so that there could one day exist a version of my display which is actually suitable
+for the above purposes.
+
+One domain where I could imagine a display like mine being especially useful is advertising.
+Because most people haven't ever seen such a display before, it would be an immediate eye-catcher
+in the display window of any business.
+
+Another possibility would be video games. Although the limited resolution and computing power
+would make it difficult to run the newest games, 3D version of classics like Pong, Snake or a
+simple platform game would be worth a shot.
+
 \end{document}

### 2016-05-17 paper.tex
fixed siunitx

@@
 \usepackage[separate-uncertainty=true, multi-part-units=single]{siunitx}
+\sisetup{load-configurations = binary}
 
@@
 Something I had in mind since the beginning of this project were animated images. The Teensy 3.1
-has \SI{256}{\kilo\byte} of memory, which would fit 42 images of \SI{6}{\kilo\byte} each. This
+has \SI{256}{kB} of memory, which would fit 42 images of \SI{6}{kB} each. This
 corresponts to roughly a second of playback. In order to display longer animations, there are
@@
 In order to get around the limit of flash memory, external storage would be necessary. Even with
-only \SI{1}{\giga\byte}, I could save 166666 frames, thus removing any practical limit in terms of
+only \SI{1}{GB}, I could save 166666 frames, thus removing any practical limit in terms of
 playback time.
@@
 Another possibility would be video games. Although the limited resolution and computing power
-would make it difficult to run the newest games, 3D version of classics like Pong, Snake or a
+would make it difficult to run the newest games, 3D versions of classics like Pong, Snake or a
 simple platform game would be worth a shot.

### 2016-05-18 paper.tex
gitignore, image creation

@@
 \usepackage[margin=2.0cm]{geometry}
+
+\usepackage{hyperref}
+
+\usepackage[]{algorithm2e}
 
@@
 
+\section{Image Creation}
+
+Initially, all my images were made completely by hand. I had to enter the individual bits into the
+program that would result in the desired image. Obviously this was a time-consuming and error
+prone task. In order to simplify the process of making an image, I wrote a python program that
+would support me in doing so. All of the source code is available on \url{github.com/mbjd/_3DPOV}.
+
+\subsection{Basics}
+
+In order to save an image, I used a three-dimensional array of 100 $\cdot$ 10 $\cdot$ 16 integers.
+The lowest three bits of each integer represent the states of the red, green and blue LED
+respectively. The rest of the bits are unused. This causes the program to use more memory than
+strictly necessary, but it makes editing the pixels much simpler.
+
+On the display itself everything stays the same, and no bits are wasted. The conversion from the
+easily editable to the memory efficient format is automatically handled by the python program.
+
+The pixels are arranged in a polar coordinate system. Because it's simpler to express straight
+lines, spheres or surfaces in terms of cartesian coordinates, calculations typically go like this:
+
+\begin{algorithm}[H]
+\ForAll{pixels in the image}{
+convert polar coordinates to cartesian\;
+calculate some function with the coordinates as input\;
+colour the pixel accordingly\;
+}
+\end{algorithm}
+
+The actual source code of the program outlined below is fully available on
+\url{http://github.com/mbjd/\_3DPOV} in the folder \texttt{image\_creation}.
+
+\subsection{Straight lines}
+
+Given two points, the program can draw a line connecting them. To do that, it calculates the
+distance to the line for each point, and if it's below a certain threshold, the point is coloured.
+This threshold controls the thickness of the line and can be selected freely, as can the colour of
+the line. The implementation can be found in the functions \texttt{drawLine}.
+
+\subsection{Spheres}
+
+Spheres are even simpler than straight lines: For each point, the program compares the distance to
+the center against the desired radius of the sphere. If it's smaller than the radius, the point is
+within the sphere and can be coloured. This functionality is implemented as \texttt{drawSphere}.
+Another version of this function, called \texttt{hollowSphere}, checks if the distance to the
+middle point is within a certain range, drawing a hollow sphere.
+
+\subsection{Planes}
+
+A plane can be defined in terms of the equation $ax + by + cz = d$, where x, y, and z are the
+cartesian coordinates, and a, b, c, and d are parameters defining the surface. Additionally, the
+function takes an argument specifying the thickness of the plane. In the end, all pixels whose
+coordinates $(x,y,z)$ satisfy $|ax + by + cz - d| < \frac{\si{thickness}}{2}$ are
+coloured with the specified colour. The implementation is called \texttt{drawSurface}.
+
+\subsection{Functions}
+
+There are two fundamental types of functions that the program can draw. The first one are functions
+of the form $f(x,y,z) \rightarrow colour$. In order to render such a function, I simply iterate
+over all pixels in the image, calculating the function and colouring the pixel with its return
+value.
+
+Secondly, the program can plot functions of the form $f(x, y) \rightarrow z$. It is a special case
+of the first one and could be defined in terms of it: I could simply iterate over all 16000 pixels
+and check if $\lfloor f(x, y) \rfloor = z$. However, that would be inefficient, because the same
+calculation would be made 10 times for the 10 different $z$ values. Instead, I only iterate over
+all combinations of $x$ and $y$ values, and colour the pixel at the corresponding $z$ value. That
+way, the function is only evaluated 1600 instead of 16000 times.
+
+The respective implementations can be found in \texttt{plotColourFunction} and
+\texttt{plotRealFunction}.
+
+There's a third one called \texttt{plotBoolFunction}. It's similar to \texttt{plotColourFunction}
+and can also be redefined in terms of it (with no significant loss in performance). Instead of
+converting the coordinates $(x, y, z)$ to a colour, it takes a function that returns a boolean
+value, and a colour. If the result of the given function is \texttt{True}, the pixel is coloured
+with the given colour, otherwise nothing happens.
+
+\subsection{Ada colourditional functionality}
+
+Using these basic building blocks, I can draw more complex shapes by repeatedly calling one of the
+above functions with different arguments. For example, there's the function
+\texttt{connectcircle}. It takes a list of points and connects each with the next one, and the
+last to the first point. That way I can draw arbitrary polygons.
+
+I can also automatically draw a cuboid by specifying two points. The cuboid is then put up between
+these points, all edges being parallel to the coordinate axes. The implementation is called
+\texttt{drawCuboid}.
+
+Another function, \texttt{connectAll}, takes a group of points and draws a line between each
+possible pair of points.  That way, I can draw a 3-sided pyramid in one go.
+
 \chapter{Operation}
@@
 
+% Somehow \bit and \byte don't work in siunitx, so I'm just using the literal strings instead
 Something I had in mind since the beginning of this project were animated images. The Teensy 3.1
@@
 
-Another possibility would be video games. Although the limited resolution and computing power
-would make it difficult to run the newest games, 3D versions of classics like Pong, Snake or a
-simple platform game would be worth a shot.
+Another fun thing to try would be video games. Although the limited resolution and computing power
+would make it difficult to run the latest and greatest games, 3D versions of classics like Pong,
+Snake or a simple platform game would certainly be worth a shot.
 

### 2016-05-18 paper.tex
Small fix
usage subsection, several small fixes

@@
 
-\usepackage{hyperref}
-
-\usepackage[]{algorithm2e}
-
+% A few useful definitions
 \usepackage{xspace}
@@
 
-% In order to include images
+% Needed to include images
 \usepackage{graphicx}
@@
 
-% To have text wrapped around images
+% To have text wrapped around figures
 \usepackage{wrapfig}
@@
 % Some parameters for wrapfig
-\setlength{\intextsep}{0pt}%
-\setlength{\columnsep}{15pt}%
-
-% Same as regular math but with a bit of space before and after
+\setlength{\intextsep}{0pt}
+\setlength{\columnsep}{15pt}
+
+% Same as regular math but with a bit of space before and after, and indented a little
 \newenvironment{ownmath}
@@
 
+\usepackage{listings} % For code snippets
+% \usepackage{courier} % For monspaced code
+\usepackage{inconsolata}
+\lstset{basicstyle=\small\ttfamily,breaklines=true}
+\usepackage[]{algorithm2e} % For pseudocode
+
 \begin{document}
@@
 
-% Introduction
 % * after chapters to not number them
-% TODO: don't show page numbers
 \chapter*{Preface}
@@
 First and foremost, I would like to express my gratitude to Mr. Patrick Spengler, my physics
-teacher and the mentor of this thesis. Not only did he solve numerous malfunctions, but he also
+teacher and the mentor of this project. Not only did he solve numerous malfunctions, but he also
 is partly responsible for my interest in electronics and engineering.
@@
 
-In the main loop of the program, this variable is continuously compared to \texttt{sinceMagnet}. If
-sinceMagnet is bigger, the LEDs are refreshed by calling \texttt{sendData}. Within that function,
-\texttt{nextPixelMicros} is incremented by \texttt{microsPerPixel}, causing the program to wait for
-the display to turn another $\frac{1}{100}$ of a full circle. After that, it'll display the next
-pixel.
+In the main loop of the program, this variable is continuously compared to \texttt{sinceMagnet}.
+If \texttt{sinceMagnet} is bigger, the LEDs are refreshed by calling \texttt{sendData}. Within
+that function, \texttt{nextPixelMicros} is incremented by \texttt{microsPerPixel}, causing the
+program to wait for the display to turn another $\frac{1}{100}$ of a full circle. After that,
+it'll display the next pixel.
 
@@
 prone task. In order to simplify the process of making an image, I wrote a python program that
-would support me in doing so. All of the source code is available on \url{github.com/mbjd/_3DPOV}.
+would support me in doing so. All of the source code is available on
+\texttt{github.com/mbjd/\_3DPOV} in the folder \texttt{image\_creation}.
 
@@
 
-The actual source code of the program outlined below is fully available on
-\url{http://github.com/mbjd/\_3DPOV} in the folder \texttt{image\_creation}.
-
 \subsection{Straight lines}
@@
 
-\subsection{Ada colourditional functionality}
+\subsection{Additional functionality}
 
@@
 possible pair of points.  That way, I can draw a 3-sided pyramid in one go.
+
+\subsection{Usage}
+
+This whole program is a python module, which is to say you can import it from any other python
+program, including the interactive python shell.
+
+\begin{lstlisting}
+$ python
+[...]
+>>> from mkmodel import *
+\end{lstlisting}
+\vspace{-2mm}
+
+After doing that, all of the functions defined in the program are available to play with. Using
+the \texttt{newImage} function, an empty image can be generated. We then draw a line and a sphere
+in the image:
+
+\begin{lstlisting}
+>>> img = newImage()
+>>> drawLinePolar(img, (0, 7, 14), (30, 2, 14), green, 0.5)
+>>> drawSpherePolar(img, (62, 4.5, 7), blue, 4.5)
+\end{lstlisting}
+\vspace{-2mm}
+
+When the desired image is created, it can be saved as an Arduino program:
+
+\begin{lstlisting}
+>>> writeSketch(img, 'example')
+\end{lstlisting}
+\vspace{-2mm}
+
+The complete program, ready to upload to the Teensy, is then saved in
+\texttt{./example/example.ino}. Creating a directory for each program is necessary because the
+Arduino IDE only opens them if they're located in a directory with the same name as the program.
 

### 2016-05-18 paper.tex
Added space after "Vdd"

@@
 the board, I still would've had to connect all the 16 anodes to \vdd. The driver chips, on the
-other hand, only need one connection to \vdd and GND each, so by placing the supply lines at the
+other hand, only need one connection to \vdd\ and GND each, so by placing the supply lines at the
 bottom of the board I have less traces running across the board vertically.

### 2016-05-23 paper.tex
Formatting, bibliography

@@
 \documentclass[a4paper, 11pt, titlepage]{report}
-
 \usepackage[margin=2.0cm]{geometry}
@@
 
+% Load the "hyperref" package, and pass the hyphens option to the url package when it is loaded by
+% hyperref. This allows urls to have \n's in them, with the correct link being clicable on both
+% lines
+\PassOptionsToPackage{hyphens}{url}\usepackage{hyperref}
+
 % Indent the first line of the first paragraph like the first lines of subsequent paragraphs
@@
 
-% Don't indent any first lines
-\setlength{\parindent}{0mm}
-
-% A little space between paragraphs
-\setlength{\parskip}{1.5mm}
+\setlength{\parindent}{0mm} % Don't indent any first lines
+\setlength{\parskip}{1.5mm} % A little space between paragraphs
 
@@
 
-As far as I know, however, only two people have built a working version of a three-dimensional
-POV display so far (and have published it on the internet).\footnote{It turned out that there was
-a third one: When I posted a picture of my project on the internet, a reddit user told me about
-the display he built years before.}
-Neither of these is being produced on a grand scale, instead they are --- like most 2D versions
---- side projects of hobbyists or engineers.
+As far as I know, however, only two people have built a working version of a three-dimensional POV
+display so far (and have published it on the internet).\footnote{It turned out that there was a
+third one: When I posted a picture of my project on the internet, a reddit user told me about the
+display he built years before.} Neither of these is being produced on a grand scale, instead they
+are --- like most 2D versions --- side projects of hobbyists or engineers.
+% todo - link or reference in footnote
 
@@
 For controlling the LEDs, I have bought a Teensy 3.1 microcontroller development board. It has a
-small footprint --- ideal for the fast rotation --- and still packs quite some power: The processor
-runs at 96 MHz at a register width of 32 bit. There are 34 configurable I/O pins.
-
-The controller has to have some kind of reference point, so that it can calculate the current speed
-and angular position of the device. For this, I used a simple hall sensor on the rotating side,
-which passes by a stationary magnet after each rotation.
+small footprint --- ideal for the fast rotation --- and still packs quite some power: The
+processor runs at 96 MHz at a register width of 32 bit. There are 34 configurable I/O pins.
+
+The controller has to have some kind of reference point, so that it can calculate the current
+speed and angular position of the device. For this, I used a simple hall sensor on the rotating
+side, which passes by a stationary magnet after each rotation.
 
@@
 
-However, the minimum data rate required to control the LEDs at an 8-bit colour depth is as follows:
+However, the minimum data rate required to control the LEDs at an 8-bit colour depth is as
+follows:
 
@@
 
-This means that I'd need at least 5 parallel \itwoc lines to control all the LEDs with 8 bit colour
-depth. Because each Teensy only has two of them, I'd need three Teensy boards, which would be
-possible but very cumbersome.
+This means that I'd need at least 5 parallel \itwoc lines to control all the LEDs with 8 bit
+colour depth. Because each Teensy only has two of them, I'd need three Teensy boards, which would
+be possible but very cumbersome.
 
@@
 Unlike most simple shift registers, the drivers' outputs are current sinks instead of current
-sources. This will be important when selecting the LEDs: In order to control each colour separately
-in an RGB LED, the cathodes have to be separated, while the anodes can be connected together.
+sources. This will be important when selecting the LEDs: In order to control each colour
+separately in an RGB LED, the cathodes have to be separated, while the anodes can be connected
+together.
 
@@
 In order for the whole device to rotate, I needed a motor. Luckily, I could salvage an appropriate
-part from an old RC motor boat. It's a big and heavy brushed motor. The fact that it has a built-in
-fan and was water-cooled in the boat leads me to think that its power output is much higher than
-what I need.
+part from an old RC motor boat. It's a big and heavy brushed motor. The fact that it has a
+built-in fan and was water-cooled in the boat leads me to think that its power output is much
+higher than what I need.
 
 Because of its high torque, I could attach the motor directly to the rotating shaft instead of
-using gears or a belt drive. That way there are less moving parts and the motor can turn at a lower
-rotational speed, both resulting in less wear and noise.
+using gears or a belt drive. That way there are less moving parts and the motor can turn at a
+lower rotational speed, both resulting in less wear and noise.
 
@@
 
-The LED drivers are arranged in a so-called daisy chain configuration, where the serial data output
-of each chip is connected to the serial data input of the next chip. This is not only the case
-within one board, but also between the different boards: The data output of one board is connected
-to the next board's input. That way, all of the 30 LED drivers can be controlled by a single SPI
-bus without anything like a chip select signal. To send data to the chips, the microcontroller
-can simply send out 480 bit without interruption.
+The LED drivers are arranged in a so-called daisy chain configuration, where the serial data
+output of each chip is connected to the serial data input of the next chip. This is not only the
+case within one board, but also between the different boards: The data output of one board is
+connected to the next board's input. That way, all of the 30 LED drivers can be controlled by a
+single SPI bus without anything like a chip select signal. To send data to the chips, the
+microcontroller can simply send out 480 bit without interruption.
 
@@
 
-The defining factor for the board layout was the pin layout of the LED drivers. The outputs are all
-at the bottom of the chip, while the data interface is at the top. This led me to the layout I have
-now: the LEDs are in a row at the bottom of the board where they can be directly connected to the
-driver outputs on the upper copper layer. The control lines are is routed on the back side of the
-board behind the chips, with vias going to the front side to connect to each chip. The data lines
-go straight from one chip to the next one on the upper copper layer, with the last output going
-back to the board's data interface on the back side. The resistors are placed on the right side of
-their respective driver IC, jumping across the data traces.
+The defining factor for the board layout was the pin layout of the LED drivers. The outputs are
+all at the bottom of the chip, while the data interface is at the top. This led me to the layout I
+have now: the LEDs are in a row at the bottom of the board where they can be directly connected to
+the driver outputs on the upper copper layer. The control lines are is routed on the back side of
+the board behind the chips, with vias going to the front side to connect to each chip. The data
+lines go straight from one chip to the next one on the upper copper layer, with the last output
+going back to the board's data interface on the back side. The resistors are placed on the right
+side of their respective driver IC, jumping across the data traces.
 
@@
 small bit of vertical board size by placing them at an offset to the right. That way I could push
-them down until where the pins begin, which is a bit further down than the edge of the chip itself.
+them down until where the pins begin, which is a bit further down than the edge of the chip
+itself.
 
@@
 With the boards finished, I wanted to know whether or not they work. In order to find out, I wrote
-a little program on the microcontroller to light the LEDs up in all possible colours consecutively.
-It did not work at the first try, so I used an oscilloscope to see if data was being sent at all.
-This was the case, the SPI library on the Teensy performed its task flawlessly.
-
-It turned out that the culprit were the bad contacts. Instead of soldering the wires onto the board
-and the microcontroller, I simply inserted them into the respective holes. This caused the
-connection to have too much parasitic capacitance, and the signals (which were being sent at 4 MHz)
-did not have time to propagate up to their destination.
+a little program on the microcontroller to light the LEDs up in all possible colours
+consecutively.  It did not work at the first try, so I used an oscilloscope to see if data was
+being sent at all.  This was the case, the SPI library on the Teensy performed its task
+flawlessly.
+
+It turned out that the culprit were the bad contacts. Instead of soldering the wires onto the
+board and the microcontroller, I simply inserted them into the respective holes. This caused the
+connection to have too much parasitic capacitance, and the signals (which were being sent at 4
+MHz) did not have time to propagate up to their destination.
 
@@
 
-In order to supply the rotating parts with electricity, I mounted a wooden disc right at the bottom
-of the shaft. On both sides of that disc I glued on a round piece of copper sheet, which I then
-connected to cables going up to where the electronics would later be.
+In order to supply the rotating parts with electricity, I mounted a wooden disc right at the
+bottom of the shaft. On both sides of that disc I glued on a round piece of copper sheet, which I
+then connected to cables going up to where the electronics would later be.
 
@@
 chamfering the top end of the shaft, the board with a \SI{4}{\milli\meter} hole just barely could
-slide on the shaft. This tight fit proved to be advantageous, because torque can now be transmitted
-reliably from the shaft to the boards, and they don't slide around.
+slide on the shaft. This tight fit proved to be advantageous, because torque can now be
+transmitted reliably from the shaft to the boards, and they don't slide around.
 
@@
 After these two additions, the angle value will go over 99 in many cases, which is the highest
-possible array index. Since the pixel at position $100+n$ is the same as the one at position $n$, I
-can pervent causing undefined behaviour by applying a modulus 100 operator to the angle value
+possible array index. Since the pixel at position $100+n$ is the same as the one at position $n$,
+I can pervent causing undefined behaviour by applying a modulus 100 operator to the angle value
 before using it to access the array.
@@
 connected the Teensy to my laptop via USB. This was dangerous because neither the Teensy nor the
-laptop's USB port can handle a lot of current, and if more than a handful of LEDs were switched on,
-the Teensy could've been damaged.
+laptop's USB port can handle a lot of current, and if more than a handful of LEDs were switched
+on, the Teensy could've been damaged.
 
@@
 
-The images displayed were constantly jittering around instead of staying still like they should. It
-took me a long time to find the source of this bug, but I managed to fix it in time for the swiss
-national competition organised by "Swiss Youth in Science".
+The images displayed were constantly jittering around instead of staying still like they should.
+It took me a long time to find the source of this bug, but I managed to fix it in time for the
+swiss national competition organised by "Swiss Youth in Science".
 
@@
 send data, control was immediately given to the former. The \texttt{timerUpdate} function was
-executed normally, resetting \texttt{currentPixel} to zero and reading out the duration of the last
-rotation. However, after that, control was given bach to the \texttt{sendData} function. At the end
-of that function, \texttt{currentPixel} was incremented by one. That way, the \texttt{currentPixel}
-variable already had a value of one when it should've stayed at 0. For the entirety of the next
-rotation, \texttt{currentPixel} was off by one, resulting in the image quickly turning by
-\SI{3.6}{\degree}.
+executed normally, resetting \texttt{currentPixel} to zero and reading out the duration of the
+last rotation. However, after that, control was given bach to the \texttt{sendData} function. At
+the end of that function, \texttt{currentPixel} was incremented by one. That way, the
+\texttt{currentPixel} variable already had a value of one when it should've stayed at 0. For the
+entirety of the next rotation, \texttt{currentPixel} was off by one, resulting in the image
+quickly turning by \SI{3.6}{\degree}.
 
@@
 would support me in doing so. All of the source code is available on
-\texttt{github.com/mbjd/\_3DPOV} in the folder \texttt{image\_creation}.
+\url{https://github.com/mbjd/\_3DPOV} in the folder \texttt{image\_creation}.
 
@@
 
+\section{Theoretical Limits}
+
+The current version of the display works fine and well, but it's certainly not at the limit of
+what can be done. In this section, I try to find out which aspects of the display can be improved
+how much.
+
+\subsection{Resolution}
+
+The resolution was determined based upon the size of the LEDs I used. Pixels are located
+\SI{3.125}{\milli\meter} apart from each other, while the LEDs themselves have a width of about
+\SI{2.8}{\milli\meter}. In order to decrease pixel size significantly, I'd have to use smaller
+pixels. To a certain extent, this would be possible; there are LEDs that are about half the width
+of the ones I use currently.
+
+With such small LEDs, there wouln't be enough space for the LED drivers. I didn't find chips with
+similar functionality in a smaller package. To a certain extent it would be possible to minimise
+the necessary space between LED drivers by using a multi-layer board layout. This method might
+just be enough for twice the pixel density, but more won't be possible. Beyond that, custom-made
+LED drivers are the only solution I can think of.
+
+There's one more approach to increasing pixel density, which is to just have one colour per
+pixel.
+
+\begin{thebibliography}{99}
+
+\bibitem{povdisp}
+\emph{POV-Display}, project report, retrieved 2016-05-19,
+\newline \url{http://www.mikrocontroller.net/articles/POV-Display}
+
+\bibitem{attiny}
+\emph{ATtiny85 POV Display}, build instructions by vishalapr, retrieved 2016-05-19
+\newline \url{http://www.instructables.com/id/ATtiny8545-POV-Display/}
+
+\bibitem{propclock}
+\emph{Build an LED Propeller Clock}, build instructions by Bob Blick, retrieved 2016-05-19
+\newline \url{http://www.jameco.com/Jameco/workshop/MyStory/diy-led-propeller-clock.html}
+
+\bibitem{volumen}
+\emph{voLumen - volumetric 3D display device}, project report by Maximilian Mali, 
+retrieved 2016-05-19
+\newline \url{https://hackaday.io/project/1737}
+
+\bibitem{hotchkiss}
+\emph{3D POV Display... Work in Progress}, video by Jason Hotchkiss, retrieved 2016-05-19
+\newline \url{https://www.youtube.com/watch?v=rlptJzWwCXg}
+
+\bibitem{povwiki}
+\emph{Persistence of Vision} on Wikipedia, retrieved 2016-05-19
+\newline \url{https://en.wikipedia.org/wiki/Persistence\_of\_vision}
+
+% http://kiel.de/index.php?action=lexikon&tag=det&id=6114 would be here too but it's not availableanymore
+
+\bibitem{teensy}
+\emph{Teensy USB Development Board}, product page, retrieved 2016-05-19
+\newline \url{https://www.pjrc.com/teensy/}
+
+\bibitem{iiclib}
+\emph{New \itwoc Library for Teensy 3}, forum post by nox771, retrieved 2016-05-19
+\newline \url{https://forum.pjrc.com/threads/21680}
+
+\bibitem{iictut}
+\emph{\itwoc Tutorial}, article, retrieved 2016-05-19
+\newline \url{http://www.best-microcontroller-projects.com/i2c-tutorial.html}
+
+\bibitem{tlc}
+\emph{TLC5927}, product page, retrieved 2016-05-19
+\newline \url{http://www.ti.com/product/TLC5927}
+
+\bibitem{tlcdigikey}
+\emph{Texas-Instruments TLC5927IDBQR}, retrieved 2016-05-19
+\newline \url{http://www.digikey.ch/product-detail/en/TLC5927IDBQR/296-24762-1-ND/}
+
+\bibitem{eagle}
+\emph{What is EAGLE}, product page, retrieved 2016-05-19
+\newline \url{http://www.cadsoftusa.com/eagle-pcb-design-software/about-eagle/}
+
+\bibitem{oshpark}
+\emph{OSH Park --- An electric ecosystem}, retrieved 2016-05-19
+\newline \url{https://oshpark.com/}
+
+\bibitem{arduino}
+\emph{Download the Arduino Software}, retrieved 2016-05-19
+\newline \url{https://www.arduino.cc/en/main/software}
+
+\bibitem{teensyduino}
+\emph{Download Teensyduino}, retrieved 2016-05-19
+\newline \url{https://www.pjrc.com/teensy/td\_download.html}
+
+\bibitem{sublime}
+\emph{Sublime Text 2}, product page, retrieved 2016-05-19
+\newline \url{http://www.sublimetext.com/2}
+
+\bibitem{stino}
+\emph{Stino}, version control page by Robot-Will, retrieved 2016-05-19
+\newline \url{https://github.com/Robot-Will/Stino}
+
+\bibitem{intervaltimer}
+\emph{IntervalTimer}, documentation page by Paul Stoffregen, retrieved 2016-05-19
+\newline \url{https://www.pjrc.com/teensy/td\_timing\_IntervalTimer.html}
+
+\bibitem{attachinterrupt}
+\emph{attachInterrput()}, documentation page, retrieved 2016-05-19
+\newline \url{https://www.arduino.cc/en/Reference/attachInterrupt}
+
+\bibitem{kickstarter}
+Comment by Paul Stoffregen, retrieved 2016-05-19
+\newline
+\url{https://www.kickstarter.com/projects/paulstoffregen/teensy-30-32-bit-arm-cortex-m4-usable-in-arduino-a/comments?cursor=1654857}
+
+\bibitem{iopins}
+\emph{Using Digital I/O Pins}, documentation page, retrieved 2016-05-19
+\newline \url{https://www.pjrc.com/teensy/td\_digital.html}
+
+\bibitem{spilib}
+\emph{SPI Library}, documentation page, retrieved 2016-05-19
+\newline \url{https://www.pjrc.com/teensy/td\_libs\_SPI.html}
+
+\bibitem{pinout}
+\emph{Teensy 3.1 Pins}, retrieved 2016-05-19
+\newline \url{https://www.pjrc.com/teensy/pinout.html}
+
+\bibitem{phoneplug}
+\emph{Phone Connector (audio)} on Wikipedia, retrieved 2016-05-19
+\newline \url{https://en.wikipedia.org/wiki/Phone\_connector}
+
+\end{thebibliography}
+
 \end{document}

### 2016-08-20 paper.tex
old updates

@@
 % todo: references
-As discussed before, there is no shortage of projects that are similar to mine, but only
-display 2D images. The name "POV display", short for persistence of vision display, has become
-commonplace. Another popular name is "Propeller Clock" --- Propeller because it rotates like one,
-and clock because the round shape is ideal for displaying an analog clock.
+As discussed before, there is no shortage of projects that are similar to mine, but only display
+2D images \cite{povdisp, attiny, propclock}. The name "POV display", short for persistence of
+vision display, has become commonplace. Another popular name is "Propeller Clock" --- Propeller
+because it rotates like one, and clock because the round shape is ideal for displaying an analog
+clock.
 
 As far as I know, however, only two people have built a working version of a three-dimensional POV
-display so far (and have published it on the internet).\footnote{It turned out that there was a
-third one: When I posted a picture of my project on the internet, a reddit user told me about the
-display he built years before.} Neither of these is being produced on a grand scale, instead they
-are --- like most 2D versions --- side projects of hobbyists or engineers.
-% todo - link or reference in footnote
+display so far and have published it online \cite{volumen, hotchkiss}.\footnote{It turned out that
+there was a third one - When I wrote a forum post about the display, somebody made a comment about
+a similar display he built years ago: \url{https://www.reddit.com/comments/2q9sg6/-/cn4i4rk}}
+Neither of these is being produced on a grand scale, instead they are --- like most 2D versions
+--- side projects of hobbyists or engineers.
 
@@
 be believed, this timespan is about $\frac{1}{25}$ of a second, but of course it is going to
-be different for each pair of eyes.
+be different for each pair of eyes \cite{povwiki}.
 
@@
 
-For controlling the LEDs, I have bought a Teensy 3.1 microcontroller development board. It has a
-small footprint --- ideal for the fast rotation --- and still packs quite some power: The
-processor runs at 96 MHz at a register width of 32 bit. There are 34 configurable I/O pins.
+For controlling the LEDs, I have bought a Teensy 3.1 microcontroller development board
+\cite{teensy}. It has a small footprint --- ideal for the fast rotation --- and still packs quite
+some power. The processor runs at 96 MHz at a register width of 32 bit, and there are 34
+configurable I/O pins.
 
@@
 
-\subsection{Data Transmission}
+\subsection{LED drivers}
 
@@
 the logic level to return to its normal state. In order to ensure correct data transmission, it is
-necessary to set a relatively slow clock speed. The two \itwoc interfaces of the Teensy only
-support clock speeds of up to 2.4 MHz --- and because \itwoc requires some overhead data to send
-out the slave's addresses, the actual data rate will be even lower.
+necessary to set a relatively slow clock speed. Even with an optimised library \cite{iiclib}, the
+two \itwoc interfaces of the Teensy are limited to 2.4 MHz if the clock speed it kept at
+\SI{96}{\mega\hertz} --- and because \itwoc requires some overhead data to send out the slave's
+addresses, the actual data rate will be even lower.
 
@@
 colour depth. Because each Teensy only has two of them, I'd need three Teensy boards, which would
-be possible but very cumbersome.
+be possible but very cumbersome. The SPI lines, however, can send a bit at every other clock
+cycle, leading to a bit rate of \SI{48}{\mega\hertz}.
 
@@
 \newpage
-\subsection{LED drivers}
 
@@
 
-After quite some searching around, I decided to use the TLC5927 by Texas Instruments. It is
-essentially a shift register with some extra features. Like every normal shift register, it has a
-serial data input (SDI), a serial data output (SDO), and a clock pin (CLK). Additionally, there is
-a latch pin (LE), which serves the purpose of only refreshing the state of the LEDs once all data
-has been sent.
+After quite some searching around, I decided to use the TLC5927 by Texas Instruments \cite{tlc,
+tlcdigikey}. It is essentially a shift register with some extra features. Like every normal shift
+register, it has a serial data input (SDI), a serial data output (SDO), and a clock pin (CLK).
+Additionally, there is a latch pin (LE), which serves the purpose of only refreshing the state of
+the LEDs once all data has been sent.
 
@@
 part from an old RC motor boat. It's a big and heavy brushed motor. The fact that it has a
-built-in fan and was water-cooled in the boat leads me to think that its power output is much
-higher than what I need.
+built-in fan and was water-cooled in the boat leads me to believe that it can take a lot more
+power than what I need for the display.
 
@@
 
-\newpage
+% \newpage
 \section{Circuit Diagram}
@@
 \begin{figure}[h]
+\vspace{0mm}
+\includegraphics[width=\textwidth]{./images/schematic.png}
+\vspace{-9mm}
+\caption{The finished circuit diagram}
 \vspace{4mm}
-\includegraphics[width=\textwidth]{./images/schematic.png}
-\vspace{-8mm}
-\caption{The finished circuit diagram}
-\vspace{6mm}
 \end{figure}
 
-The circuit diagram --- as well as the PCB layout --- were created in Eagle. Its free version is
-limited to two-layer boards of sizes up to \SI{10}{\centi\meter} by \SI{8}{\centi\meter}, but
-that's more than enough for my purposes.
+The circuit diagram --- as well as the PCB layout --- were created in Eagle \cite{eagle}. Its free
+version is limited to two-layer boards of sizes up to \SI{10}{\centi\meter} by
+\SI{8}{\centi\meter}, but that's more than enough for my purposes.
 
@@
 
+\subsection{General Remarks}
+
+The defining factor for the board layout was the pin layout of the LED drivers. The outputs are
+all at the bottom of the chip, while the data interface is at the top. This led me to the layout I
+have now: the LEDs are in a row at the bottom of the board where they can be directly connected to
+the driver outputs on the upper copper layer. The control lines are routed on the back side of
+the board behind the chips, with vias going to the front side to connect to each chip. The data
+lines go straight from one chip to the next one on the upper copper layer, with the last output
+going back to the board's data interface on the back side. The resistors are placed on the right
+side of their respective driver IC, jumping across the data traces.
+
 \begin{figure}[h]
@@
 green = vias, grey = silkscreen.}
-\vspace{6mm}
+\vspace{-3mm}
 \end{figure}
-
-\subsection{General Remarks}
-
-The defining factor for the board layout was the pin layout of the LED drivers. The outputs are
-all at the bottom of the chip, while the data interface is at the top. This led me to the layout I
-have now: the LEDs are in a row at the bottom of the board where they can be directly connected to
-the driver outputs on the upper copper layer. The control lines are is routed on the back side of
-the board behind the chips, with vias going to the front side to connect to each chip. The data
-lines go straight from one chip to the next one on the upper copper layer, with the last output
-going back to the board's data interface on the back side. The resistors are placed on the right
-side of their respective driver IC, jumping across the data traces.
 
@@
 providers to ship small production runs within a reasonable timeframe and with good production
-quality turned out to be OSHPark. They only produce in multiples of 3, so I ordered 12 instead of
-10 (which I'd have done anyway to have some spare boards). I submitted my order on the 5th of
-October 2014 and paid US\$ 47. Ten days later I received a message that the boards were finished,
-and after another ten days of waiting they arrived at my doorstep.
+quality turned out to be OSHPark \cite{oshpark}. They only produce in multiples of 3, so I ordered
+12 instead of 10 (which I'd have done anyway to have some spare boards). I submitted my order on
+the 5th of October 2014 and paid US\$ 47. Ten days later I received a message that the boards were
+finished, and after another ten days of waiting they arrived at my doorstep.
 
@@
 
-The following program has been compiled with the Arduino IDE 1.0.6 and sent to the Teensy by the
-Teensyduino 1.20 plugin. However, the most part of the program has been written in Sublime Text 2
-with the stino plugin. This combination has a few advantages over simply using the Arduino IDE to
-write all code, namely there's automatic code completion, support for several cursors, and a
-search-and-replace function with regex suport. I installed and used these programs on Mac OS 10.9
-and later 10.10.
+The following program has been compiled with the Arduino IDE 1.0.6 \cite{arduino} and sent to the
+Teensy by the Teensyduino 1.20 plugin \cite{teensyduino}. However, the most part of the program
+has been written in Sublime Text 2 \cite{sublime} with the stino plugin \cite{stino}. This
+combination has a few advantages over simply using the Arduino IDE to write all code, namely
+there's automatic code completion, support for several cursors, and a search-and-replace function
+with regex suport. I installed and used these programs on Mac OS 10.9 and later 10.10.
 
@@
 that function, \texttt{nextPixelMicros} is incremented by \texttt{microsPerPixel}, causing the
-program to wait for the display to turn another $\frac{1}{100}$ of a full circle. After that,
-it'll display the next pixel.
-
+program to wait for the display to turn another $\frac{1}{100}$ of a full circle. After that, the
+next pixel will be displayed.
+
+% TODO put source code in appendix
 The implementation of the described behaviour can be found in the appendix, where the whole source
-code is to be found. Specifically, what I've just talked about is contained in the
+code is included. Specifically, what I've just talked about is contained in the
 \texttt{timerUpdate} function, as well as in the main \texttt{loop}.
@@
 each byte on a row of LEDs. Because there are 48 LEDs per row, 6 bytes can be used to describe the
-row's state without wasting any storage space. The only way to be more efficient in terms of
-storage would be to compresss data, which would require significantly more computing power.
-
-Because the PCBs are not right above each other, the data that is being sent out in one refresh
+row's state without wasting any storage space.
+
+Because the PCBs are not right above each other, the data that is being sent out at the same time
 don't correspond to the pixels that are above each other in the image. In order avoid having to
@@
 
-\bibitem{iictut}
-\emph{\itwoc Tutorial}, article, retrieved 2016-05-19
-\newline \url{http://www.best-microcontroller-projects.com/i2c-tutorial.html}
-
 \bibitem{tlc}
@@
 
-\bibitem{intervaltimer}
-\emph{IntervalTimer}, documentation page by Paul Stoffregen, retrieved 2016-05-19
-\newline \url{https://www.pjrc.com/teensy/td\_timing\_IntervalTimer.html}
-
-\bibitem{attachinterrupt}
-\emph{attachInterrput()}, documentation page, retrieved 2016-05-19
-\newline \url{https://www.arduino.cc/en/Reference/attachInterrupt}
-
-\bibitem{kickstarter}
-Comment by Paul Stoffregen, retrieved 2016-05-19
-\newline
-\url{https://www.kickstarter.com/projects/paulstoffregen/teensy-30-32-bit-arm-cortex-m4-usable-in-arduino-a/comments?cursor=1654857}
+% \bibitem{intervaltimer}
+% \emph{IntervalTimer}, documentation page by Paul Stoffregen, retrieved 2016-05-19
+% \newline \url{https://www.pjrc.com/teensy/td\_timing\_IntervalTimer.html}
+
+% \bibitem{attachinterrupt}
+% \emph{attachInterrput()}, documentation page, retrieved 2016-05-19
+% \newline \url{https://www.arduino.cc/en/Reference/attachInterrupt}
+
+% \bibitem{kickstarter}
+% Comment by Paul Stoffregen, retrieved 2016-05-19
+% \url{https://www.kickstarter.com/projects/paulstoffregen/teensy-30-32-bit-arm-cortex-m4-usable-in-arduino-a/comments?cursor=1654857}
 

### 2016-08-20 paper-short.tex
moved shorter version and summaries to this repo

\documentclass[a4paper, 11pt, titlepage]{article}
\usepackage[margin=2.0cm]{geometry}

% A few useful definitions
\usepackage{xspace}
\def \itwoc {I\textsuperscript{2}C\xspace}
\def \vdd {V\textsubscript{DD}}

\linespread{1.35}

% Load the "hyperref" package, and pass the hyphens option to the url package when it is loaded by
% hyperref. This allows urls to have \n's in them, with the correct link being clicable on both
% lines
\PassOptionsToPackage{hyphens}{url}\usepackage{hyperref}

% \usepackage{indentfirst}
\setlength{\parindent}{0mm} % Don't indent any first lines
\setlength{\parskip}{1mm} % A little space between paragraphs

% Needed to include images as figures
\usepackage{graphicx}

% Label figures 1, 2, 3, 4... Instead of 1.1, 1.2, 2.1, 3.1 etc.
\usepackage{chngcntr}
\counterwithout{figure}{section}

% Display SI units
\usepackage[binary-units=true, separate-uncertainty=true, multi-part-units=single]{siunitx}

% To have text wrapped around figures
\usepackage{wrapfig}

% Some parameters for wrapfig
\setlength{\intextsep}{0pt}
\setlength{\columnsep}{15pt}

% Same as regular math but with a bit of space before and after, and indented a little
\newenvironment{ownmath}
{\vspace{2mm}\hspace{15pt}\begin{math}}
{\end{math}\vspace{2mm}}

\begin{document}

\begin{titlepage}

    \centering
    \vspace*{3cm}
    {\huge\bfseries Development of a 3D Display}
    \rule{0.67\textwidth}{1.5pt}
    \vspace*{2cm}

    % \hspace*{-1cm}
    \centerline{\includegraphics[width=0.9\paperwidth]{./images/cover.jpg}}

    \vspace*{1cm}
    \Large
    Balduin Dettling

    \large
    Matura Project 2014 \\ Kantonsschule Wiedikon

    Mentored by Patrick Spengler and Prof. Heinz Domeisen
    \vfill

\end{titlepage}

\section{Introduction}

\pagenumbering{arabic}

3D Displays are becoming increasingly common: Movies can be viewed in 3D cinemas, and people even
have 3D television screens in their living room. However, these displays don't actually display
depth, they merely provide an illusion of it. They work by projecting images from slightly
different viewpoints to each eye, letting the human brain figure out depth information.

In this work, however, I built a display that is actually three-dimensional. It can be viewed from
any direction, and pixels are distributed across three spatial dimensions. Such a display is
called a persistence of vision display or a swept-volume display. It uses a rapidly rotating
two-dimensional arrangement of LEDs. By controlling it precisely, it can display a still
three-dimensional image.

The idea to build such a display came from two-dimensional displays building upon the same
principle; building them is quite a popular project among tinkerers and engineers alike. I
initially had in mind to make another 2D display, but since there are instruction manuals online I
didn't see much of a point in doing so. Instead I went one step further and built a
three-dimensional version.

In my display, 160 pixels are distributed across ten circuit boards. All LEDs are controlled by a
Teensy 3.1 microcontroller and 30 LED drivers. They are refreshed 100 times per rotation, for a
total of 16000 voxels. The display rotates at about \SI{60}{\hertz}, which is enough for the image
to not flicker. By measuring the rotational speed with a hall effect sensor, the image stays in
place despite the rotational speed not being exactly constant.

A python program facilitates the creation of images: It can insert straight lines, spheres,
cuboids, planes and even plots of arbitrary functions into the an image, which can then be loaded
onto the Teensy and shown on the 3D display.

\section{Hardware}

\subsection{Components}

\subsubsection{Control}

For controlling the LEDs, I used a Teensy 3.1 microcontroller development board. Its small
footprint is ideal for the fast rotation. With a clock speed \SI{96}{\mega\hertz} and a register
width of \SI{32}{\bit}, it's powerful enough to rapidly control a large amount of LEDs.

The controller has to have some kind of reference point, so that it can calculate the current
speed and angular position of the device. For this, I used a hall effect sensor on the rotating
side, which passes by a stationary magnet after each rotation.

This microcontroller can't control all 480 LEDs on its own, which is why I used LED drivers.
Specifically, I used the TLC5927 by Texas Instruments. These chips essentially consist of a
\SI{16}{\bit} shift register with some extra features, the most important of them being constant
current regulation.

\subsubsection{LEDs}

The LEDs I used are surface-mounted RGB leds with a diffused lens and a width of
\SI{2.8}{\milli\meter}. Their small size makes them ideal to use as pixels, and the diffused lens
causes the colors mix well, even when viewed directly.

\subsubsection{Motor}

In order to make the whole device rotate, I salvaged a brushed DC motor from an old RC motor boat.
Because of its high torque, I could attach the motor directly to the rotating shaft instead of
using gears or a belt drive. That way there are less moving parts and the motor can turn at a
lower rotational speed, both resulting in less wear and noise.

\subsubsection{Power Transmission}

Power is transmitted from the stationary power supply to the rotating electronics by a pair of
slip rings. I made them out of two round pieces of copper sheet on the rotating end, and two long
pieces of copper sheet on the stationary part.

In order to supply 480 LEDs with \SI{20}{\milli\ampere} each, a current of \SI{9.6}{\ampere} is
necessary. Testing has shown that the slip rings don't cause a significant voltage drop at this
current. In practice, however, the average current is much lower since most pixels are switched
off in a typical 3D image.

Because two pieces of copper rubbing against each other aren't the most reliable electrical
connection, I used a \SI{2200}{\micro\farad} electrolytic capacitor to smooth out the supply
voltage.

\subsection{Circuit Diagram}

\begin{figure}[h]
\vspace{0mm}
\includegraphics[width=\textwidth]{./images/schematic.png}
\vspace{-10mm}
\caption{The finished circuit diagram}
\vspace{4mm}
\end{figure}

The circuit diagram --- as well as the PCB layout --- were created in Eagle. Its free version is
limited to two-layer boards of sizes up to \SI{10}{\centi\meter} by \SI{8}{\centi\meter}, but
that's more than enough for my purposes.

The LED drivers are arranged in a daisy chain configuration, where the serial data output of each
chip is connected to the serial data input of the next chip. This is not only the case within one
board, but also between the different boards: The data output of one board is connected to the
next board's input. That way, all 30 LED drivers can be controlled by a single SPI bus.

\def \iout {I\textsubscript{OUT}\xspace}
\def \rext {R\textsubscript{EXT}\xspace}

The external resistor connected to each LED driver sets the output current. The following
equation, taken from page 15 in the datasheet, describes the relation between the output current
(\iout) and the resistor's value (\rext):

% Lots of ugly hacks to have a non-italic font
\begin{ownmath}
\textrm{\iout} = 15 \cdot \frac{\SI{1.25}{\volt}}{\si{\rext}}
\end{ownmath}

% \frac{\SI{1.25}{\volt}}{\SI{\rext}}
For an output current of \SI{20}{\milli\ampere}, \rext needs to be \SI{937.5}{\ohm}. The next
smaller value within the E24 series is \SI{910}{\ohm}, which is what I ultimately used.

A \SI{1}{\micro\farad} bypass capacitor is located close to each LED driver so that their power
consumption can change almost instantly without causing a spike in the supply voltage.

I used a row of 5 pins to transmit data to the LED drivers. The data lines (serial data in
and out) are connected to the input of the first chip and the output of the last chip
respectively. The rest of the pins (clock, output enable and latch enable) are control lines,
which go to all three chips simultaneously.

\subsection{PCB Design}

The above schematic translates directly to the PCB layout, which was also created in Eagle. The
most important consideration was to save as much space as possible, which I did simply by placing
parts as close as possible to each other. The supply lines are routed on top of each other so
their magnetic fields (which can change quickly) are largely cancelled out.

\begin{figure}[h]
\vspace{4mm}
\includegraphics[width=\textwidth]{./images/board-layout.png}
\vspace{-8mm}
\caption{The PCB layout in Eagle.}
\vspace{-5mm}
\end{figure}

\section{Fabrication}

\subsection{Circuit Boards}

I ordered 12 PCBs with the above layout at \url{https://oshpark.com/}, which has cost me 47 USD.
20 days later, the finished boards arrived at my doorstep.

\begin{figure}[h]
\vspace{3mm}
\includegraphics[width=\textwidth]{./images/board-finished.jpeg}
\vspace{-10mm}
\caption{A finished circuit board}
\vspace{2mm}
\end{figure}

Soldering the tiny, surface-mount LEDs to the board took some practice, but otherwise I
encountered no problems while equipping the boards with components.

\subsection{Physical Assembly}

The frame is built out of beech wood strips on a plywood plate. A horizontal board hosts the
motor, which in turn supports the rotating shaft with all the electronic on it.

Around the frame, I constructed an enclosure out of acrylic glass to protect the display and to
remove the risk of accidentally reaching into the rotating circuit boards.

The slip rings used to transmit power to the rotating electronics consist of two round pieces of
copper sheet on the rotating part, and two long pieces of copper sheet on the stationary part.
Unfortunately, they need a considerable amount of pressure to form a good enough contact, which
leads to quick wear as well as wasted electricity.

The microcontroller is mounted on another wooden disc, located a few centimetres above the slip
rings. The \SI{2200}{\micro\farad} capacitor as well as the hall sensor are also located on this
disc. The corresponding magnet is mounted on a piece of threaded rod, which itself is mounted on
the same board as the motor.

In the end, this place has become a bit of a mess, especially with all the cables going up to the
circuit boards. Luckily this isn't visible while the device is running.

Finally, I mounted the circuit boards on the shaft connected to the motor. There were two
important factors I had to consider while arranging the boards:

\begin{itemize}
\item The boards had to be balanced out to minimise static and dynamic unbalance
\item The amount of pixels obstructed by other circuit boards had to be minimal
\end{itemize}

I fulfilled both of these requirements by arranging the ten boards in a double-spiral arrangement.
This was the only arrangement I could think of that (theoretically) has no static imbalance at all
and minimal dynamic imbalance. On the viewer-facing half of the display, all LEDs are constantly
visible. On the backside, however, it's possible that one group of boards is right before the
other one, causing some layers of the image to disappear. When viewing the display from above,
almost all pixels except some of the innermost ones are visible.

\section{Programming}

\subsection{Timing}

In order to display a stable image, the program has to compensate for slight changes in rotational
speed. The hall effect sensor is used to measure the duration of one revolution. After this
measurement is made, the value is divided by 100 and stored in the variable
\texttt{microsPerPixel}, where it's interpreted as a reference for how long each pixel should be
displayed.

A variable called \texttt{nextPixelMicros} stores the time when the next pixel should be
displayed. As soon as this time has arrived, the function \texttt{sendData} is called and
\texttt{nextPixelMicros} is incremented by \texttt{microsPerPixel}, causing the program to wait
until the next pixel should be dispayed.

\subsection{Data storage and display}

The image is saved in a three-dimensional array of $100 \cdot 10 \cdot 6$ bytes. The first index
stands for the angle of a pixel in hundredths of a circle, the second index for the height and the
third one for the position of each byte on a row of LEDs. Because there are 48 LEDs per row, 6
bytes can be used to describe the row's state without wasting any storage space.

The previously mentioned function \texttt{sendData} needs to be able to know which bytes to send
based on the current angle of of the display. Because the PCBs are not right above each other, the
bytes that are being sent out at the same time don't correspond to the pixels that are above each
other in the image. The program accounts for these offsets by calculating the position of each
board relative to the first one and sending the according values out of the saved image.

\subsection{Image Creation}

Initially, I created all the 3D images by hand, which was a tedious and error prone task. In order
to simplify it, I wrote a python program that would support me in making images. All of the source
code is available on \url{https://github.com/mbjd/\_3DPOV} in the folder \texttt{image\_creation}.

In order to save an image in this program, I used a three-dimensional array of $100 \cdot 10 \cdot
16$ integers.  The lowest three bits of each integer represent the states of the red, green and
blue LED respectively. In comparison to the data format used on the display itself, this takes up
more memory but is much simpler to edit.

The program can create simple shapes such as straight lines, planes, spheres, cuboids or arbitrary
polygons. It can also plot any given function of the signature $f(x,y) \mapsto z$, $f(x,y,z)
\mapsto \text{colour}$, or $f(x,y,z) \mapsto \{0, 1\} $, where x, y, and z are the three axes of
the cartesian coordinate system.

Once an image is complete, it can be converted to a program for the Teensy 3.1 with a single
function call.

\section{Conclusions}

\subsection{Possible Improvements}

The display works as initially planned and can display volumetric images with a resolution of $16
\cdot 10 \cdot 100$ pixels and a colour depth of \SI{3}{\bit}. It's also something that most
people have never seen, which makes them interested in the display. However, there are several
aspects that could be improved upon.

\subsubsection{Slip Rings}

Because the slip rings are quite rough, substantial pressure is needed to form a consistent
contact. This leads to a lot of friction and quick wear of the contacts.

This problem could be solved by replacing the slip rings with two coils for inductive power
transfer. While this would be a more complicated solution, it would significantly reduce friction
and noise while transmitting power more reliably.

\subsubsection{Image creation}

The program I wrote facilitates creating simple images with geometric shapes. However, the only
way to make more complicated images is to set each pixel by hand. A more flexible solution would
be a program that takes a 3D model in a widely used file format and converts it to be displayable
on my display, so users can create their images in any 3D graphics program.

\subsubsection{Memory Space}

Since the Teensy 3.1 only has \SI{256}{\kilo\byte} of flash memory, the amount of images that can
be stored is very limited. For static images this is fine, but animations can have a length of
only 42 frames before running out of space.

In order to get around this limit, external storage would be necessary, for example in the form of
an SD card. Even with \SI{1}{\giga\byte}, I could save 166666 frames, thus removing any practical
limit in terms of playback time.

\subsubsection{Interactive Programs}

It would be interesting to not only display predefined images, but to actually run a program that
can respond to user input. In order to communicate with the external world, a wireless interface
would be necessary. Readily available solutions include small WLAN or bluetooth modules, or the
transmitters and receivers commonly used in radio-controlled aircraft.

\subsection{Applications}

Because of the low resolution and colour depth, the high level of noise, quick wear and the
inability to send images to the display while running, this display is not suited as a replacement
for a computer or TV screen. However, these problems can all be fixed completely or partially, so
that there could one day exist a version of my display which is actually suitable for the above
purposes.

One domain where I could imagine a display like mine being especially useful is advertising.
Because most people haven't ever seen such a display before, it would be an immediate eye-catcher
in the display window of any business.

Furthermore, a display like mine would certainly be useful in educational settings. By visualising
things like atomic orbitals, the earth's structure, orbital mechanics or meteorologic phenomena on
a 3D display rather than on paper, these topics could be understood more easily.

Another fun thing to try would be video games. Although the limited resolution and computing power
would make it difficult to run the latest and greatest games, 3D versions of classics like Pong,
Snake or a simple platform game would certainly be worth a shot.

\end{document}

### 2016-08-20 straightforward.tex
moved shorter version and summaries to this repo

\documentclass[a4paper, 11pt]{article}
\usepackage[margin=2cm]{geometry}
\linespread{1.5}
\setlength{\parindent}{0mm}
\setlength{\parskip}{2mm}

\pagenumbering{gobble}

\begin{document}

\hfill Balduin Dettling

\vspace{1cm}
\begin{center}

	{\huge\bfseries Development of a 3D Display\vspace{-2mm}}
	\rule{0.67\textwidth}{1.5pt}

	{\vspace{-2mm} \Large\bfseries Straightforward Description}

\end{center}

In this project, I have built a three-dimensional persistence of vision
display. Unlike most 3D displays (like television or cinema screens), it's a
\emph{volumetric} display, which means that it actually displays depth instead
of merely providing an illusion of it.

It works by quickly rotating 160 RGB LEDs. A microcontroller controls them such
that they draw an image into the air. This happens so fast that the human eye
doesn't see quick bursts of light, but steadily lit pixels.

\end{document}

### 2016-08-20 summary.tex
moved shorter version and summaries to this repo

\documentclass[a4paper, 11pt]{article}
\usepackage[margin=2cm]{geometry}
\linespread{1.5}
\setlength{\parindent}{0mm}
\setlength{\parskip}{2mm}

\usepackage{siunitx}

\pagenumbering{gobble}

\begin{document}

\hfill Balduin Dettling

\vspace{1cm}
\begin{center}

    {\huge\bfseries Development of a 3D Display\vspace{-2mm}} 
    \rule{0.67\textwidth}{1.5pt}

    {\vspace{-2mm} \Large\bfseries Summary}

\end{center}

\section*{Goals}

The aim of this project was to build a working prototype of a volumetric persistence of vision
display.

\section*{Methods}

The display consists of a two-dimensional arrangement of LEDs that rotates at about
\SI{60}{\hertz}. A microcontroller controls the LEDs such that they produce a volumetric image
that can be viewed from any direction. Using a hall effect sensor and a magnet, the display
measures its rotational speed and adjusts its timing accordingly, so that the image stays in place
despite the rotational frequency slightly varying.

Because the rotational period is smaller than the flicker fusion threshold of the human eye, the
image appears to be steady and the observer sees the whole image at all times, even though at any
given time only $\frac{1}{100}$ of the pixels are illuminated.

A python program makes it easy to insert simple geometric shapes or plots of a user-given function
into an image, and it can convert the generated images to programs for the microcontroller that
display the image.

\section*{Conclusions}

The initial goal has been reached and the display works as planned. Possible applications include
advertising, education and video games. My display could be improved by incorporating wireless
data transmission, so that images can be changed at runtime, or by replacing the loud and
inefficient slip rings with inductive power transfer. Another worthwile addition would be a
program that converts 3D models in a popular format to the format used by my display, which is
nothing but a 3D bitmap.

\end{document}

### 2017-01-14 paper.tex
use \frac{...}{...} instead of \si{ ... \per ...}
fixed typo, used siunitx more

@@
 \item Colour depth of 3 bit per pixel, or one bit per LED
-\item Rotational frequency of at least 30 Hz, or as fast as it needs to be in order to not flicker
+\item Rotational frequency of at least \SI{30}{\hertz}, or as fast as it needs to be in order
+	to not flicker
 
@@
 \cite{teensy}. It has a small footprint --- ideal for the fast rotation --- and still packs quite
-some power. The processor runs at 96 MHz at a register width of 32 bit, and there are 34
-configurable I/O pins.
+some power. The processor runs at \SI{96}{\mega\hertz} at a register width of 32 bit, and there
+are 34 configurable I/O pins.
 
@@
 most LED drivers modulate at frequencies within that range. In my case, however, the LEDs are
-moving into a new pixel 3000 times a second (assuming rotationat 30 Hz). This means that a PWM
+moving into a new pixel 3000 times a second (assuming rotation at 30 Hz). This means that a PWM
 frequency under 3000 Hz will involve losing information, and one slightly above 3000 Hz still
 won't look good. There would have to be a significant number of PWM cycles within each pixel,
-for example 10, which would require a PWM frequency of 30 KHz.
+for example 10, which would require a PWM frequency of \SI{30}{\kilo\hertz}.
 
@@
 so I decided to place the LEDs at \SI{3}{\milli\metre} increments. On a grid of \SI{12.5}{mil},
-the next larger distance is \SI{125}{mil}, which is equal to \SI{3.175}{mm}.
+the next larger distance is \SI{125}{mil}, which is equal to \SI{3.175}{\milli\metre}.
 
@@
 \textrm{P\textsubscript{MAX}} = (\SI{5}{\volt} - \textrm{V\textsubscript{F}}) \cdot
-\SI{20.6}{\milli\ampere\per output} \cdot \SI{16}{outputs} = \SI{780}{\milli\watt}
+\num{20.6} \frac{\si{\milli\ampere}}{{\text{output}}} \cdot \SI{16}{outputs} = \SI{780}{\milli\watt}
 \end{ownmath}
@@
 board and the microcontroller, I simply inserted them into the respective holes. This caused the
-connection to have too much parasitic capacitance, and the signals (which were being sent at 4
-MHz) did not have time to propagate up to their destination.
+connection to have too much parasitic capacitance, and the signals (which were being sent at
+\SI{4}{\mega\hertz}) did not have time to propagate up to their destination.
 
